---
title: "County-Level Wildfire Risk Forecasting Under Asymmetric Error Costs: Logistic Regression vs. XGBoost"
author: "Zahir Sabbah"
date: "March 16, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
    pandoc_args: ["--quiet"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

```{css, echo=FALSE}
/* Keep floating TOC always on top and give it a solid background */
#TOC {
  position: fixed;
  z-index: 1000;
  background-color: #ffffff;
  border-right: 1px solid #e0e0e0;
  padding: 10px;
  overflow-y: auto;
  max-height: 90vh;
}

/* Prevent main content from sliding under the TOC */
.col-md-9 {
  position: relative;
  z-index: 1;
}
```

---

# Introduction

Wildfires represent one of the most destructive and costly natural hazards in the United States. Between 1992 and 2015, over 1.88 million fires were recorded across the country, burning tens of millions of acres and costing billions of dollars in suppression, property damage, and ecological loss. The western United States has been disproportionately affected, but fire risk is not static — frequency and severity have trended upward over recent decades as prolonged drought, earlier snowmelt, and shifting vegetation patterns create increasingly favorable conditions for large fire events.

Predicting where and when notable wildfires will occur is operationally valuable: fire agencies must make resource allocation decisions weeks in advance of fire season, and a model that can flag high-risk county-months with meaningful accuracy could improve pre-positioning of crews, aircraft, and equipment. This report approaches that prediction problem as a binary classification task, asking: given the fire history and geographic context of a U.S. county, what is the probability that a notable wildfire (≥300 acres) will occur there in a given month?

---

# Abstract

We develop a predictive model for the monthly occurrence of notable wildfires (≥300 acres) across U.S. counties. Using 1.88 million wildfire records aggregated at the county-month level, we engineer temporal, spatial, and lagged features and compare logistic regression to an XGBoost classifier. A time-based training/testing split (train: 1992–2012, test: 2013–2015) is used to avoid data leakage and simulate real-world forecasting conditions. We argue that AUC — while useful for threshold-independent discrimination comparison — is not the appropriate primary metric for hazard prediction tasks where false negatives and false positives carry asymmetric costs. A model that misses a fire leaves agencies unprepared; a false alarm results in unnecessary but recoverable pre-positioning. Under this cost structure, recall is the operationally decisive metric, and we evaluate both models on this basis. XGBoost achieves substantially higher recall than logistic regression through deliberate imbalance correction, while maintaining comparable AUC — a distinction that matters for deployment even when the aggregate discrimination gap appears modest. Summer seasonality and prior-month fire activity emerge as the strongest predictors across both models.

---

# Section 1: Data Preprocessing

The raw wildfire dataset contains 1.88 million individual fire records spanning 1992–2015. Before modeling, individual fire events must be filtered, spatially and temporally aggregated into structured observation units, and enriched with engineered features that capture the dynamics known to drive wildfire risk.

## 1.1 Loading the Data

```{r load-data, results='hide'}
library(DBI)
library(RSQLite)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
library(pROC)
library(knitr)
library(kableExtra)
library(maps)
library(viridis)
library(broom)
library(scales)
library(patchwork)
library(plotly)
library(leaflet)
library(htmltools)

# Connect to database — update path if needed
con <- dbConnect(SQLite(), "~/Downloads/FPA_FOD_20170508.sqlite")
fires_raw <- dbReadTable(con, "Fires")
dbDisconnect(con)

cat("Raw dataset dimensions:", nrow(fires_raw), "rows,", ncol(fires_raw), "columns\n")
```

## 1.2 Defining Notable Fires

A wildfire is classified as **notable** if it burns ≥300 acres. This threshold aligns with federal reporting standards distinguishing large fires from routine vegetation burns, and ensures the outcome variable captures events with meaningful ecological and economic impact.

The raw dataset contains fires of all sizes, from single-acre brush fires to multi-hundred-thousand-acre conflagrations. Modeling all recorded fires would create extreme class imbalance and would conflate routine small fires — often suppressed within hours — with the large, persistent events that drive suppression costs, evacuations, and ecological damage. The 300-acre cutoff is consistent with operational definitions used by U.S. fire management agencies, where fires in grass or brush fuel models are classified as "large" at this threshold. Fires below this size are excluded from the outcome variable but their aggregated counts still inform predictor construction (e.g., `roll3_fires`).

```{r filter-fires}
fires_clean <- fires_raw %>%
  filter(!is.na(FIRE_SIZE), !is.na(FIPS_CODE), !is.na(FIRE_YEAR)) %>%
  mutate(
    notable = as.integer(FIRE_SIZE >= 300),
    FIPS_CODE = as.character(FIPS_CODE)
  )

total_fires     <- nrow(fires_raw)
after_filter    <- nrow(fires_clean)
notable_count   <- sum(fires_clean$notable)
pct_notable     <- round(100 * notable_count / after_filter, 2)
pct_retained    <- round(100 * after_filter / total_fires, 2)

filter_summary <- data.frame(
  Metric = c("Total raw fire records",
             "Records after removing missing FIPS/year (36% dropped)",
             "Notable fires (≥300 acres)",
             "% of cleaned records that are notable",
             "% of raw records retained"),
  Value = c(
    format(total_fires, big.mark = ","),
    format(after_filter, big.mark = ","),
    format(notable_count, big.mark = ","),
    paste0(pct_notable, "%"),
    paste0(pct_retained, "%")
  )
)

kable(filter_summary, caption = "Table 1: Fire Filtering Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## 1.3 Constructing Space–Time Units (County × Month)

Each observation in the modeling dataset represents a **county-month pair**: a unique combination of U.S. county (identified by FIPS code) and calendar month. This spatial resolution balances granularity with statistical reliability — finer units (e.g., grid cells) would introduce excessive sparsity, while coarser units (e.g., state-year) would lose predictive resolution.

The county is a natural spatial unit for this problem for several reasons. County boundaries are stable over the study period, align with how fire agencies organize jurisdictions and resources, and are small enough to capture meaningful geographic variation in climate and vegetation while large enough to accumulate sufficient fire observations. The month is the natural temporal unit given that wildfire risk varies strongly by season, and monthly prediction horizons are operationally useful for resource pre-positioning.

Once fires are assigned to county-month cells, we aggregate by taking the maximum of the `notable` indicator within each cell — so a county-month is coded as 1 if *any* notable fire occurred, regardless of how many.

```{r construct-units, results='hide'}
fires_clean <- fires_clean %>%
  mutate(
    DISCOVERY_DATE_REAL = as.Date("1899-12-30") + DISCOVERY_DATE,
    MONTH = month(DISCOVERY_DATE_REAL),
    YEAR  = FIRE_YEAR,
    STATE = trimws(STATE),
    FIPS  = paste0(STATE, "_", FIPS_CODE)
  )

# Aggregate to county-month level
county_month <- fires_clean %>%
  group_by(FIPS, STATE, YEAR, MONTH) %>%
  summarise(
    notable_fire    = as.integer(any(notable == 1)),
    n_fires         = n(),
    total_area      = sum(FIRE_SIZE, na.rm = TRUE),
    .groups = "drop"
  )

cat("County-month observations:", nrow(county_month), "\n")
cat("Unique counties:", n_distinct(county_month$FIPS), "\n")
cat("Years covered:", min(county_month$YEAR), "to", max(county_month$YEAR), "\n")
```

## 1.4 Defining the Outcome Variable

The binary outcome is formally defined as:

$$Y_{i,t} = \begin{cases} 1 & \text{if } \geq 1 \text{ notable wildfire occurred in county } i \text{ during month } t \\ 0 & \text{otherwise} \end{cases}$$

where $i$ indexes U.S. counties and $t$ indexes calendar months.

This binary formulation is appropriate for several reasons. First, the operational question — "will a notable fire occur?" — is inherently binary. Second, modeling fire count or burned area introduces additional complexity (zero-inflation, heavy tails) that is unnecessary given our forecasting goal. Third, binary classification allows straightforward evaluation via AUC and the ROC framework, which are well-suited to imbalanced outcomes.

It is important to note that this formulation treats a county-month with one small notable fire the same as one with a catastrophic multi-thousand-acre event. This is an intentional simplification: the goal is occurrence prediction, not severity prediction. Future work could extend this to an ordinal or continuous outcome.

## 1.5 Predictor Construction

We engineer the following predictors motivated by domain knowledge of wildfire dynamics. Rather than using raw columns directly, each predictor is designed to capture a specific aspect of fire risk: seasonality, long-term trend, geographic context, and temporal autocorrelation.

**Temporal features** (`MONTH_F`, `SEASON`, `YEAR`): Wildfire risk is strongly seasonal, driven by temperature, precipitation, and vegetation dryness cycles. Month is encoded as a 12-level factor to allow each month its own baseline risk, while `SEASON` provides a coarser 4-level grouping used in the XGBoost model. `YEAR` captures any long-term secular trend in fire frequency — important given documented increases in wildfire activity over recent decades linked to drought and climate change.

**Spatial features** (`REGION`): U.S. Census regions capture broad climate and vegetation differences that systematically drive fire risk. The West — with its Mediterranean climate, dry summers, and chaparral and conifer forests — faces fundamentally different fire dynamics than the humid Southeast or the Midwest. Region acts as a fixed effect, absorbing baseline risk differences that cannot be explained by temporal variables alone.

**Lag features** (`lag1_fire`, `lag1_area`, `roll3_fires`): Fire activity exhibits strong temporal autocorrelation — a county that experienced a notable fire last month is substantially more likely to experience one this month. This pattern is counterintuitive at first glance: a fire burns through fuel, so one might expect the landscape to be depleted and a subsequent fire unlikely for years afterward. The data tell the opposite story, and the explanation lies in the conditions that cause fires rather than the fires themselves.

Large wildfires are not random events — they require a specific convergence of conditions: prolonged drought that desiccates vegetation, high temperatures, low humidity, and wind. These conditions are driven by large-scale atmospheric patterns — heat domes, persistent high-pressure ridges, La Niña drought cycles — that do not reset after a single fire. When a county burns in July, it is almost certainly because that entire region is in the grip of exceptional heat and drought. Those same conditions persist into August and September. Adjacent fuel beds, unburned areas within the same county, and areas where the fire was suppressed before burning completely are all still exposed to the same atmospheric forcing. A fire in month $t$ is therefore a strong signal that the *underlying conditions* driving fire risk are still present in month $t+1$ — not that the same ground will re-ignite, but that the same drought and heat that caused the first fire have not abated.

Secondary mechanisms reinforce this: large fires produce embers that travel miles and ignite new spot fires; fire-weakened trees fall and create additional fuel loads; and suppression resources stretched by one large fire may be slower to respond to ignitions elsewhere in the same region. `lag1_fire` is a binary indicator for notable fire in the prior month; `lag1_area` captures the magnitude of prior-month burning (log-transformed to reduce skewness); and `roll3_fires` counts total fires in the preceding three months as a measure of sustained regional fire activity.

**Climate features** (`pdsi`, `tmpc`, `pcpn`): Three climate division variables from NOAA's nClimDiv dataset are added as XGBoost-only features. All three are published in the same fixed-width format and downloaded from the same NOAA directory.

- **PDSI** (Palmer Drought Severity Index): measures cumulative soil moisture deficit. Negative values indicate drought. Captures multi-month drying that precedes large fire events.
- **Temperature** (`tmpc`): mean monthly temperature in °F by climate division. High temperatures accelerate fuel drying and raise ignition probability directly.
- **Precipitation** (`pcpn`): monthly total precipitation in inches. Low precipitation months deplete live fuel moisture and allow dead fuels to dry toward ignition thresholds.

Together these three variables capture the full atmospheric state driving fire risk: how dry the soil is (PDSI), how hot it is (tmpc), and how much recent moisture has fallen (pcpn). All three have nonlinear relationships with fire occurrence — thresholds matter more than linear slopes — which is precisely why they are reserved for XGBoost rather than logistic regression. The three data files are included in this repository; the code below detects them automatically.

<details>
<summary>**Data source** (click to expand)</summary>
Files downloaded from NOAA's nClimDiv archive: `https://www.ncei.noaa.gov/pub/data/cirs/climdiv/`  
Filenames: `climdiv-pdsidv-v1.0.0-YYYYMMDD`, `climdiv-tmpcdv-v1.0.0-YYYYMMDD`, `climdiv-pcpndv-v1.0.0-YYYYMMDD`  
Saved as `climdiv-pdsidv.txt`, `climdiv-tmpc.txt`, `climdiv-pcpn.txt` in the repository root. If absent the model runs without climate features.
</details>

```{r climdiv-integration, results='hide'}
# ── NOAA nClimDiv Integration ──────────────────────────────────────────────────
# Files: climdiv-pdsidv.txt, climdiv-tmpc.txt, climdiv-pcpn.txt
# Download from: https://www.ncei.noaa.gov/pub/data/cirs/climdiv/

noaa_state_map <- c(
  "01"="AL","02"="AZ","03"="AR","04"="CA","05"="CO","06"="CT",
  "07"="DE","08"="FL","09"="GA","10"="ID","11"="IL","12"="IN",
  "13"="IA","14"="KS","15"="KY","16"="LA","17"="ME","18"="MD",
  "19"="MA","20"="MI","21"="MN","22"="MS","23"="MO","24"="MT",
  "25"="NE","26"="NV","27"="NH","28"="NJ","29"="NM","30"="NY",
  "31"="NC","32"="ND","33"="OH","34"="OK","35"="OR","36"="PA",
  "37"="RI","38"="SC","39"="SD","40"="TN","41"="TX","42"="UT",
  "43"="VT","44"="VA","45"="WA","46"="WV","47"="WI","48"="WY"
)

parse_climdiv <- function(filepath, varname) {
  raw    <- readLines(filepath)
  parsed <- do.call(rbind, lapply(raw, function(line) {
    parts <- strsplit(trimws(line), "\\s+")[[1]]
    if (length(parts) < 13) return(NULL)
    header <- parts[1]
    year   <- as.integer(substr(header, 7, 10))
    if (is.na(year) || year < 1000) return(NULL)
    STATE  <- noaa_state_map[substr(header, 1, 2)]
    if (is.na(STATE)) return(NULL)
    vals <- suppressWarnings(as.numeric(parts[2:13]))
    data.frame(STATE = STATE, YEAR = year, MONTH = 1:12,
               val = vals, stringsAsFactors = FALSE)
  }))
  if (is.null(parsed) || nrow(parsed) == 0) return(NULL)
  parsed <- parsed[!is.na(parsed$val) & parsed$val > -99, ]
  parsed %>% group_by(STATE, YEAR, MONTH) %>%
    summarise(!!varname := mean(val, na.rm = TRUE), .groups = "drop")
}

find_climdiv <- function(prefix) {
  for (nm in c(paste0(prefix, ".txt"), paste0(prefix, ".csv"), prefix)) {
    if (file.exists(nm)) return(nm)
  }
  hits <- list.files(".", pattern = paste0("^", prefix), full.names = TRUE,
                     ignore.case = TRUE)
  if (length(hits) > 0) return(hits[1])
  return(NULL)
}

pdsi_file <- find_climdiv("climdiv-pdsidv")
tmpc_file <- find_climdiv("climdiv-tmpc")
pcpn_file <- find_climdiv("climdiv-pcpn")

climdiv_available <- !is.null(pdsi_file) || !is.null(tmpc_file) || !is.null(pcpn_file)
climdiv_data      <- NULL

if (!is.null(pdsi_file)) {
  d <- parse_climdiv(pdsi_file, "pdsi")
  if (!is.null(d)) climdiv_data <- if (is.null(climdiv_data)) d else
    full_join(climdiv_data, d, by = c("STATE","YEAR","MONTH"))
}
if (!is.null(tmpc_file)) {
  d <- parse_climdiv(tmpc_file, "tmpc")
  if (!is.null(d)) climdiv_data <- if (is.null(climdiv_data)) d else
    full_join(climdiv_data, d, by = c("STATE","YEAR","MONTH"))
}
if (!is.null(pcpn_file)) {
  d <- parse_climdiv(pcpn_file, "pcpn")
  if (!is.null(d)) climdiv_data <- if (is.null(climdiv_data)) d else
    full_join(climdiv_data, d, by = c("STATE","YEAR","MONTH"))
}
```

```{r feature-engineering, results='hide'}
# Sort for lag computation
county_month <- county_month %>%
  arrange(FIPS, YEAR, MONTH)

# Lag features: prior month fire activity
county_month <- county_month %>%
  group_by(FIPS) %>%
  mutate(
    lag1_fire       = lag(notable_fire, 1),
    lag1_area       = lag(total_area, 1),
    roll3_fires     = lag(n_fires, 1) + lag(n_fires, 2) + lag(n_fires, 3)
  ) %>%
  ungroup()

# Add region variable
region_map <- tibble::tribble(
  ~STATE, ~REGION,
  "WA","West",  "OR","West",  "CA","West",  "ID","West",  "MT","West",
  "WY","West",  "NV","West",  "UT","West",  "CO","West",  "AZ","West",
  "NM","West",  "AK","West",  "HI","West",
  "ND","Midwest","SD","Midwest","NE","Midwest","KS","Midwest",
  "MN","Midwest","IA","Midwest","MO","Midwest","WI","Midwest",
  "MI","Midwest","IL","Midwest","IN","Midwest","OH","Midwest",
  "TX","South",  "OK","South",  "AR","South",  "LA","South",
  "MS","South",  "AL","South",  "TN","South",  "KY","South",
  "WV","South",  "VA","South",  "NC","South",  "SC","South",
  "GA","South",  "FL","South",
  "ME","Northeast","NH","Northeast","VT","Northeast","MA","Northeast",
  "RI","Northeast","CT","Northeast","NY","Northeast","NJ","Northeast",
  "PA","Northeast","MD","Northeast","DE","Northeast"
)

county_month <- county_month %>%
  left_join(region_map, by = "STATE") %>%
  mutate(
    REGION = ifelse(is.na(REGION), "West", REGION),
    MONTH_F  = factor(MONTH, levels = 1:12,
                      labels = c("Jan","Feb","Mar","Apr","May","Jun",
                                 "Jul","Aug","Sep","Oct","Nov","Dec")),
    SEASON = case_when(
      MONTH %in% c(12, 1, 2) ~ "Winter",
      MONTH %in% c(3, 4, 5)  ~ "Spring",
      MONTH %in% c(6, 7, 8)  ~ "Summer",
      TRUE                    ~ "Fall"
    )
  )

# Remove rows with NAs from lagging
model_data <- county_month %>%
  filter(!is.na(lag1_fire), !is.na(lag1_area), !is.na(roll3_fires), !is.na(REGION)) %>%
  mutate(notable_fire = factor(notable_fire, levels = c(0, 1)))

# ── Inline nClimDiv join ──────────────────────────────────────────────────────
# Read and join each climate file directly here, avoiding cross-chunk scoping issues

noaa_map <- c("01"="AL","02"="AZ","03"="AR","04"="CA","05"="CO","06"="CT",
              "07"="DE","08"="FL","09"="GA","10"="ID","11"="IL","12"="IN",
              "13"="IA","14"="KS","15"="KY","16"="LA","17"="ME","18"="MD",
              "19"="MA","20"="MI","21"="MN","22"="MS","23"="MO","24"="MT",
              "25"="NE","26"="NV","27"="NH","28"="NJ","29"="NM","30"="NY",
              "31"="NC","32"="ND","33"="OH","34"="OK","35"="OR","36"="PA",
              "37"="RI","38"="SC","39"="SD","40"="TN","41"="TX","42"="UT",
              "43"="VT","44"="VA","45"="WA","46"="WV","47"="WI","48"="WY")

read_climdiv_inline <- function(filepath, varname) {
  raw <- readLines(filepath)
  rows <- do.call(rbind, lapply(raw, function(line) {
    p <- strsplit(trimws(line), "\\s+")[[1]]
    if (length(p) < 13) return(NULL)
    h <- p[1]; yr <- as.integer(substr(h, 7, 10))
    if (is.na(yr) || yr < 1000) return(NULL)
    st <- noaa_map[substr(h, 1, 2)]
    if (is.na(st)) return(NULL)
    vals <- suppressWarnings(as.numeric(p[2:13]))
    data.frame(STATE=st, YEAR=yr, MONTH=1:12, v=vals, stringsAsFactors=FALSE)
  }))
  if (is.null(rows) || nrow(rows) == 0) return(NULL)
  rows <- rows[!is.na(rows$v) & rows$v > -99, ]
  rows %>% group_by(STATE, YEAR, MONTH) %>%
    summarise(!!varname := mean(v, na.rm=TRUE), .groups="drop")
}

for (cv_info in list(
    list(prefix="climdiv-pdsidv", var="pdsi"),
    list(prefix="climdiv-tmpc",   var="tmpc"),
    list(prefix="climdiv-pcpn",   var="pcpn")
)) {
  f <- NULL
  for (nm in c(paste0(cv_info$prefix, ".txt"), paste0(cv_info$prefix, ".csv"),
               cv_info$prefix)) {
    if (file.exists(nm)) { f <- nm; break }
  }
  if (is.null(f)) {
    hits <- list.files(".", pattern=paste0("^", cv_info$prefix), full.names=TRUE,
                       ignore.case=TRUE)
    if (length(hits) > 0) f <- hits[1]
  }

  if (!is.null(f)) {
    df <- read_climdiv_inline(f, cv_info$var)
    if (!is.null(df)) {
      model_data <- model_data %>%
        left_join(df, by = c("STATE", "YEAR", "MONTH"))
      n_miss <- sum(is.na(model_data[[cv_info$var]]))
      tr_mean <- mean(model_data[[cv_info$var]][model_data$YEAR <= 2012], na.rm=TRUE)
      if (n_miss > 0) model_data[[cv_info$var]] <-
        ifelse(is.na(model_data[[cv_info$var]]), tr_mean, model_data[[cv_info$var]])
    } else {
      model_data[[cv_info$var]] <- 0
    }
  } else {
    model_data[[cv_info$var]] <- 0
  }
}

cat("Final modeling dataset:", nrow(model_data), "observations\n")
cat("Positive outcome rate:", round(100 * mean(as.integer(model_data$notable_fire) - 1), 2), "%\n")
```

```{r feature-table}
feature_table <- data.frame(
  Variable       = c("notable_fire", "MONTH_F", "YEAR", "REGION",
                     "lag1_fire", "lag1_area", "roll3_fires", "SEASON",
                     "pdsi", "tmpc", "pcpn"),
  Description    = c("Binary outcome: ≥1 notable fire in county-month",
                     "Calendar month (factor, 12 levels)",
                     "Calendar year (numeric trend)",
                     "U.S. Census region (factor)",
                     "Notable fire in same county previous month",
                     "Total burned area in same county previous month",
                     "Rolling 3-month fire count in same county",
                     "Season of year (factor)",
                     "Palmer Drought Severity Index (NOAA nClimDiv, state-month)",
                     "Mean temperature °F (NOAA nClimDiv, state-month)",
                     "Total precipitation inches (NOAA nClimDiv, state-month)"),
  Type           = c("Binary (0/1)", "Factor", "Numeric", "Factor",
                     "Binary (0/1)", "Numeric", "Numeric", "Factor",
                     "Numeric", "Numeric", "Numeric"),
  Motivation     = c("Outcome", "Captures seasonality",
                     "Captures long-term trend", "Captures regional climate differences",
                     "Temporal autocorrelation", "Severity persistence",
                     "Sustained activity window", "Simplified seasonal signal",
                     "Cumulative soil moisture deficit — multi-month drought signal",
                     "Heat driving fuel drying and ignition probability",
                     "Recent moisture suppressing or enabling fire conditions")
)

kable(feature_table, caption = "Table 2: Predictor Variables") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE)
```

```{r dataset-summary-table}
# Required professor output: dataset-level summary statistics
n_obs       <- nrow(model_data)
n_counties  <- n_distinct(model_data$FIPS)
n_months    <- n_distinct(paste(model_data$YEAR, model_data$MONTH))
n_positive  <- sum(as.integer(model_data$notable_fire) - 1)
pct_pos     <- round(100 * n_positive / n_obs, 2)
yr_range    <- paste(min(model_data$YEAR), "–", max(model_data$YEAR))

dataset_summary <- data.frame(
  Characteristic = c(
    "Total county-month observations",
    "Unique counties (FIPS units)",
    "Unique year-month combinations",
    "Years covered",
    "County-months with notable fire (Y = 1)",
    "County-months without notable fire (Y = 0)",
    "Positive outcome rate"
  ),
  Value = c(
    format(n_obs,      big.mark = ","),
    format(n_counties, big.mark = ","),
    format(n_months,   big.mark = ","),
    yr_range,
    format(n_positive,        big.mark = ","),
    format(n_obs - n_positive, big.mark = ","),
    paste0(pct_pos, "%")
  )
)

kable(dataset_summary,
      caption = "Table 3: Final Modeling Dataset Summary") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

## 1.6 Missing Data Handling

```{r missing-data, results='hide'}
n_missing_firesize <- sum(is.na(fires_raw$FIRE_SIZE))
n_missing_fips     <- sum(is.na(fires_raw$FIPS_CODE))
n_missing_year     <- sum(is.na(fires_raw$FIRE_YEAR))
n_dropped_missing  <- nrow(fires_raw) - nrow(fires_clean)
pct_dropped        <- round(100 * n_dropped_missing / nrow(fires_raw), 2)

n_lag_dropped <- sum(!complete.cases(county_month[, c("lag1_fire","lag1_area","roll3_fires")]))
```

```{r missing-data-table}
missing_summary <- data.frame(
  Source = c("Missing FIRE_SIZE", "Missing FIPS_CODE", "Missing FIRE_YEAR",
             "Total records dropped (missing)", "County-months dropped (lag NAs)"),
  Count  = c(format(n_missing_firesize, big.mark=","),
             format(n_missing_fips,     big.mark=","),
             format(n_missing_year,     big.mark=","),
             paste0(format(n_dropped_missing, big.mark=","),
                    " (", pct_dropped, "% of raw)"),
             format(n_lag_dropped, big.mark=","))
)

kable(missing_summary, caption = "Table 4: Missing Data Summary",
      col.names = c("Source", "Records Affected")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

Records with missing `FIRE_SIZE`, `FIPS_CODE`, or `FIRE_YEAR` were excluded prior to aggregation. As Table 4 shows, the dominant source of missingness is `FIPS_CODE`: approximately 678,000 records (~36% of the raw dataset) were dropped on this basis. The missingness in FIPS code likely reflects jurisdictions where county-level coding was incomplete rather than a random data quality issue. The remaining ~1.2 million records are distributed across 2,795 counties and 24 years, retaining strong national coverage. However, the model predicts fire risk only for counties with some presence in the database — counties systematically absent due to FIPS missingness are not represented, which may introduce geographic selection bias toward jurisdictions with more complete administrative records.

To assess whether FIPS missingness is geographically patterned, we compute the proportion of dropped records by state and compare it to each state's share of retained records.

```{r fips-missingness-geo}
missing_by_state <- fires_raw %>%
  mutate(missing_fips = is.na(FIPS_CODE)) %>%
  group_by(STATE) %>%
  summarise(
    total      = n(),
    n_missing  = sum(missing_fips),
    pct_missing = round(100 * n_missing / total, 1),
    .groups = "drop"
  ) %>%
  filter(n_missing > 0) %>%
  arrange(desc(pct_missing)) %>%
  slice_head(n = 12) %>%
  select(STATE, total, n_missing, pct_missing) %>%
  rename(`State` = STATE,
         `Total Records` = total,
         `Missing FIPS` = n_missing,
         `% Missing` = pct_missing)

kable(missing_by_state,
      caption = "Table 4b: States with Highest FIPS Missingness Rate (Top 12)",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped","condensed"), full_width = FALSE)
```

**Interpretation:** FIPS missingness is not randomly distributed across states. Several states with very high missingness rates — DC, Puerto Rico, Alaska, the Dakotas — are either non-contiguous, small, or low-fire-risk jurisdictions whose exclusion has limited effect on the modeling dataset's coverage of fire-prone regions. However, Table 4b also shows that California (70.3% missing), Arizona (65.7%), and Montana (60.7%) — all high-fire-risk western states — appear prominently among the top missingness states. This means a meaningful share of the dropped records originates from the very regions where fire risk is highest. The practical implication is that the modeling dataset likely underrepresents some high-intensity fire events in these states, which could cause the model to underestimate risk in the most fire-prone western counties. This is a genuine selection bias that cannot be fully resolved without the missing FIPS codes. It is acknowledged as a limitation of the data rather than the modeling approach, and the retained dataset still contains substantial fire observations from western states — sufficient for the models to identify the regional and seasonal patterns documented in Section 2 — but the bias should be kept in mind when interpreting absolute risk estimates for high-activity western counties.

Lag variables introduce NAs for the first recorded observation per county (since there is no prior month to lag from). These rows are dropped from the final modeling dataset — approximately 8,161 county-months as shown in Table 4. No imputation was performed, as this missingness is entirely structural — it is a consequence of the lag construction, not a data quality problem — and imputing the first observation with zeros would artificially suppress fire probability estimates for counties' earliest records.

---

# Section 2: Exploratory Data Analysis

Four structural features of the data — severe class imbalance, nonlinear seasonality, a region-season interaction, and threshold-shaped lag predictors — each have direct implications for model selection.

## 2.1 Class Imbalance in the Outcome Variable

Only 4.43% of county-months contain a notable fire. A classifier trained without accounting for this imbalance will learn to predict zero almost everywhere, achieving high accuracy while being operationally useless — which is why accuracy is not reported as a primary metric.

```{r class-imbalance-table}
class_counts <- model_data %>%
  mutate(Outcome = ifelse(as.integer(notable_fire) - 1 == 1,
                          "Notable Fire (Y = 1)", "No Notable Fire (Y = 0)")) %>%
  count(Outcome) %>%
  mutate(Percentage = paste0(round(100 * n / sum(n), 2), "%"),
         n = format(n, big.mark = ",")) %>%
  rename(`County-Month Observations` = n)

kable(class_counts,
      caption = "Table 5: Class Distribution of the Binary Outcome Variable",
      col.names = c("Outcome", "Count", "% of Dataset")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

**Interpretation:** Only 4.43% of county-months contain a notable fire. A classifier that always predicts "no fire" would achieve ~95.6% accuracy while being operationally useless, which is why accuracy is not reported as a primary metric. Standard logistic regression on heavily imbalanced data is pulled toward predicting low probabilities for almost every observation. XGBoost addresses this directly via `scale_pos_weight`, which up-weights fire months during training without altering the test distribution.

## 2.2 Temporal Pattern of Notable Wildfire Occurrence

```{r plot-temporal, fig.cap="Figure 1: Monthly rate of notable wildfire occurrence across all county-month observations, 1992–2015. Each point represents the proportion of county-months in that calendar month with at least one notable fire."}
temporal_plot <- model_data %>%
  group_by(YEAR, MONTH) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop")

# Correct average: mean of per-year monthly rates (equal weight per year)
monthly_avg <- temporal_plot %>%
  group_by(MONTH) %>%
  summarise(avg_rate = mean(fire_rate), .groups = "drop")

month_labels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

temporal_plot <- temporal_plot %>%
  mutate(month_label = month_labels[MONTH],
         tooltip_text = paste0("Year: ", YEAR,
                               "<br>Month: ", month_labels[MONTH],
                               "<br>Fire Rate: ", scales::percent(fire_rate, accuracy = 0.1)))

monthly_avg <- monthly_avg %>%
  mutate(month_label = month_labels[MONTH],
         tooltip_text = paste0("Cross-year Average<br>Month: ", month_labels[MONTH],
                               "<br>Avg Fire Rate: ", scales::percent(avg_rate, accuracy = 0.1)))

p_temp <- ggplot(temporal_plot,
                 aes(x = MONTH, y = fire_rate, group = YEAR, color = YEAR,
                     text = tooltip_text)) +
  geom_line(alpha = 0.4, linewidth = 0.6) +
  geom_line(data = monthly_avg,
            aes(x = MONTH, y = avg_rate, group = 1, text = tooltip_text),
            color = "#d62728", linewidth = 1.5, inherit.aes = FALSE) +
  scale_color_viridis_c(option = "plasma", name = "Year") +
  scale_x_continuous(breaks = 1:12, labels = month_labels) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    title = "Monthly Notable Wildfire Rate by Year (1992–2015)",
    x     = "Month",
    y     = "Proportion of Counties with Notable Fire"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())

ggplotly(p_temp, tooltip = "text") %>%
  layout(
    height = 450,
    margin = list(t = 70, r = 20, b = 60, l = 70),
    annotations = list(
      list(x = 0, y = 1.08, text = "Red line = average across all years | Hover lines to explore by year",
           xref = "paper", yref = "paper", showarrow = FALSE,
           font = list(size = 11, color = "grey40"))
    )
  )
```

**Interpretation:** Wildfire occurrence exhibits pronounced seasonality. Individual years show their highest rates in the July–September window, with the cross-year average peaking in September — likely reflecting the outsized weight of western counties, where fire seasons extend later into fall than in other regions. The average rate rises from near zero in winter to a peak of approximately `r { temporal_plot <- model_data %>% group_by(MONTH) %>% summarise(fire_rate = mean(as.integer(notable_fire)-1), .groups="drop"); paste0(round(100*max(temporal_plot$fire_rate), 1), "%") }` of county-months, a roughly 5-fold increase over the spring baseline. The spike is abrupt: risk climbs steeply from June onward and collapses after October. Logistic regression encodes month as 12 additive log-odds shifts, which can approximate this shape but imposes smoothness the data does not support. Gradient boosted trees place a split directly on the summer–fall threshold without that constraint.

The September peak in the national average is a hypothesis worth verifying: if it is driven by the West's late fire season, we should see the West alone peak in September while other regions peak earlier. The figure below confirms this directly.

```{r temporal-by-region, fig.cap="Figure 1b: Cross-year average monthly fire rate broken out by U.S. Census region. The West peaks in September, consistent with its late-summer/fall fire season; the South peaks in spring (May); the Midwest shows a secondary spring peak before dropping through summer; the Northeast remains near zero throughout. This confirms that the September peak in the national average (Figure 1) is an artefact of western county weighting — not a national pattern.", fig.width=10, fig.height=5}
regional_monthly_avg <- model_data %>%
  group_by(REGION, YEAR, MONTH) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  group_by(REGION, MONTH) %>%
  summarise(avg_rate = mean(fire_rate), .groups = "drop")

region_colors <- c(West = "#d62728", South = "#e07b00",
                   Midwest = "#1f77b4", Northeast = "#2ca02c")

month_labels_short <- c("J","F","M","A","M","J","J","A","S","O","N","D")

ggplot(regional_monthly_avg, aes(x = MONTH, y = avg_rate, color = REGION, group = REGION)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = region_colors, name = NULL) +
  scale_x_continuous(breaks = 1:12, labels = month_labels_short) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  facet_wrap(~REGION, nrow = 1) +
  labs(
    title   = "Monthly Fire Rate by Region: Confirming the September Peak is Western",
    x       = "Month",
    y       = "Avg. Proportion of Counties with Notable Fire"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "none",
        panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"),
        axis.text.x      = element_text(size = 9))
```

This regional decomposition confirms the hypothesis and has a direct modeling implication: the "season" effect in the national data is not uniform — it is a West-specific fall effect superimposed on a South-specific spring effect. The Midwest also shows a modest spring peak before declining through summer, while the Northeast remains near zero throughout. A model that treats season as a global additive term systematically misestimates risk for all regions. The `month_region` interaction feature in XGBoost's feature matrix is designed precisely to capture this structure.

## 2.3 Spatial Pattern of Wildfire Risk by State

```{r plot-spatial, fig.cap="Figure 2: Average monthly probability of notable wildfire occurrence by U.S. state. States with no data are shown in grey. Hover to see state name and fire rate."}
state_risk <- model_data %>%
  group_by(STATE) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  mutate(STATE_lower = tolower(state.name[match(STATE, state.abb)]),
         STATE_full  = state.name[match(STATE, state.abb)])

# Use built-in state boundary data via maps package
us_states <- maps::map("state", fill = TRUE, plot = FALSE)
state_ids  <- sapply(strsplit(us_states$names, ":"), `[`, 1)

# Build lookup from cleaned state names to fire rate
rate_lookup <- setNames(state_risk$fire_rate, state_risk$STATE_lower)
# unname() prevents NA-named elements from breaking htmlwidgets serialisation
state_rates <- unname(rate_lookup[state_ids])

pal <- colorNumeric(
  palette  = "YlOrRd",
  domain   = c(0, max(state_risk$fire_rate, na.rm = TRUE)),
  na.color = "#cccccc"
)

# Tooltip: state name + formatted rate
state_full_names <- tools::toTitleCase(state_ids)
tooltip_labels <- unname(ifelse(
  is.na(state_rates),
  paste0("<b>", state_full_names, "</b><br>No data"),
  paste0("<b>", state_full_names, "</b><br>Avg monthly fire rate: ",
         scales::percent(state_rates, accuracy = 0.01))
))

leaflet(data = us_states, width = "100%", height = 480) %>%
  addProviderTiles(providers$CartoDB.PositronNoLabels) %>%
  addPolygons(
    fillColor   = pal(state_rates),
    fillOpacity = 0.85,
    color       = "white",
    weight      = 1,
    label       = lapply(tooltip_labels, htmltools::HTML),
    labelOptions = labelOptions(
      style     = list("font-weight" = "normal", padding = "4px 8px"),
      textsize  = "13px",
      direction = "auto"
    ),
    highlightOptions = highlightOptions(
      weight       = 2.5,
      color        = "#333",
      fillOpacity  = 1,
      bringToFront = TRUE
    )
  ) %>%
  addLegend(
    pal      = pal,
    values   = state_rates,
    position = "bottomright",
    title    = "Avg Monthly<br>Fire Rate",
    labFormat = labelFormat(transform = function(x) round(x * 100, 1),
                            suffix = "%")
  ) %>%
  setView(lng = -96, lat = 38, zoom = 4)
```

**Interpretation:** Wildfire risk is heavily concentrated in the western United States. Western states reach average monthly fire rates of several percent per county-month, while the Midwest and Northeast are substantially lower — with the Midwest showing seasonal elevation in spring but averaging well below western rates annually. The highest-risk western states show average rates approximately `r { west_rate <- model_data %>% filter(REGION == "West") %>% summarise(r = round(100 * mean(as.integer(notable_fire)-1), 1)) %>% pull(r); ne_rate <- model_data %>% filter(REGION == "Northeast") %>% summarise(r = round(100 * mean(as.integer(notable_fire)-1), 2)) %>% pull(r); paste0(round(west_rate / max(ne_rate, 0.01)), "×") }` higher than those in the Northeast. The deeper implication — that geographic effects cannot be separated from seasonal ones — is examined in Section 2.4.

There is a critical caveat to this map: the states that appear darkest — California (CA), Arizona (AZ), and Montana (MT) — are precisely the states identified in Table 4b as having the highest FIPS code missingness rates (70.3%, 68.9%, and 64.3% respectively). Because fires without a valid FIPS code are excluded from the modeling dataset, notable fires in these states are systematically undercounted. The map therefore *understates* true western fire risk, particularly for California. The implication for model validity is that the West's coefficient in the logistic regression and the region feature in XGBoost are both calibrated on an incomplete picture of the most fire-prone region — a structural limitation that weather covariates with complete spatial coverage (such as the nClimDiv temperature and precipitation variables) could partially compensate for.

## 2.4 Fire Rate by Season and Region

```{r plot-season-region, fig.cap="Figure 3: Average notable wildfire rate by season and U.S. region. Lines connect the four seasons for each region, making cross-region differences and within-region seasonal trajectories simultaneously visible."}
season_region <- model_data %>%
  group_by(REGION, SEASON) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  mutate(SEASON = factor(SEASON, levels = c("Winter","Spring","Summer","Fall")))

region_colors <- c(West = "#d62728", South = "#e07b00",
                   Midwest = "#1f77b4", Northeast = "#2ca02c")

ggplot(season_region, aes(x = SEASON, y = fire_rate,
                          color = REGION, group = REGION)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  geom_text(aes(label = scales::percent(fire_rate, accuracy = 0.1)),
            vjust = -1, size = 3.2, fontface = "bold", show.legend = FALSE) +
  scale_color_manual(values = region_colors, name = "Region") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0.05, 0.18))) +
  labs(
    title = "Notable Wildfire Rate by Season and Region",
    x     = "Season",
    y     = "Proportion of County-Months with Notable Fire"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "bottom")
```

**Interpretation:** The interaction between season and region is the most important finding for model selection. The West experiences its peak fire activity in Summer, the South peaks in Spring, and the Midwest shows a secondary Spring elevation before dropping — only the Northeast remains near zero in every season. Risk is not additive: the West in Summer is categorically more dangerous than either factor alone predicts, and the South in Spring departs entirely from what an additive model would expect.

The South's spring peak warrants a deeper look. *Why* does the South peak in spring rather than summer? This pattern is well-documented in fire ecology: the southeastern United States has a long tradition of prescribed burning — intentional, managed fires used to maintain longleaf pine ecosystems, improve wildlife habitat, and reduce fuel loads. The prescribed burn season in the Southeast runs predominantly from late winter through late spring (February–May), when relative humidity is moderate, temperatures are cool enough to control fire spread, and vegetation is dormant enough to minimise smoke impacts on populated areas. The data's spring peak in the South therefore reflects a deliberate management practice rather than uncontrolled wildfire risk driven by drought and heat. This distinction matters for model interpretation: a high predicted probability for a South county in May is partially capturing the presence of managed burning, not just hazardous conditions.

```{r south-spring-detail, fig.cap="Figure 3b: Monthly fire rate for the South region alone, shown month-by-month. The May peak is consistent with the southeastern prescribed burn season (February–May), a deliberate land management practice rather than a drought-driven hazard signal.", fig.width=7, fig.height=4}
south_monthly <- model_data %>%
  filter(REGION == "South") %>%
  group_by(YEAR, MONTH) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  group_by(MONTH) %>%
  summarise(avg_rate = mean(fire_rate), .groups = "drop")

ggplot(south_monthly, aes(x = MONTH, y = avg_rate)) +
  geom_col(aes(fill = (MONTH %in% 2:5)), width = 0.7, show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "#aec7e8", "TRUE" = "#e07b00")) +
  annotate("rect", xmin = 1.5, xmax = 5.5, ymin = 0, ymax = Inf,
           alpha = 0.08, fill = "#e07b00") +
  annotate("text", x = 3.5, y = max(south_monthly$avg_rate) * 0.92,
           label = "Prescribed\nburn season", color = "#7a4400",
           size = 3.2, fontface = "italic") +
  scale_x_continuous(breaks = 1:12,
                     labels = c("Jan","Feb","Mar","Apr","May","Jun",
                                "Jul","Aug","Sep","Oct","Nov","Dec")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "South Region: Monthly Fire Rate",
       subtitle = "Orange = prescribed burn season (Feb–May)",
       x = "Month", y = "Avg. Fire Rate") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

Capturing this interaction requires either manually specifying region × month interaction terms in logistic regression — 44 additional parameters that increase overfitting risk substantially — or using a model that discovers interactions through tree structure. XGBoost takes the latter approach.

## 2.5 Predictor Distributions: Fire vs. Non-Fire Observations

```{r plot-categorical-predictors, fig.cap="Figure 4a: Fire rate within each predictor category, with the dashed line marking the overall dataset baseline of 4.4%. A bar above the baseline indicates that group raises fire probability; below means it lowers it."}

baseline_rate <- mean(as.integer(model_data$notable_fire) - 1)

lag1_summary <- model_data %>%
  mutate(lag1_raw = as.character(lag1_fire)) %>%
  group_by(lag1_raw) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1),
            n = n(), .groups = "drop") %>%
  mutate(
    pct_of_data = round(100 * n / sum(n)),
    lag1_label  = ifelse(lag1_raw == "1",
                         paste0("Prior Fire\n(~", pct_of_data, "% of months)"),
                         paste0("No Prior Fire\n(~", pct_of_data, "% of months)")),
    lag1_label  = factor(lag1_label, levels = lag1_label[order(fire_rate)])
  )

p_lag1 <- ggplot(lag1_summary, aes(x = lag1_label, y = fire_rate, fill = fire_rate)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_hline(yintercept = baseline_rate, linetype = "dashed",
             color = "grey40", linewidth = 0.7) +
  geom_text(aes(label = scales::percent(fire_rate, accuracy = 0.1)),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_gradient(low = "#1f77b4", high = "#d62728") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.20))) +
  labs(title = "Prior Month Fire", x = NULL, y = "Fire Rate") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())

p_region <- model_data %>%
  group_by(REGION) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  arrange(desc(fire_rate)) %>%
  mutate(REGION = factor(REGION, levels = REGION)) %>%
  ggplot(aes(x = REGION, y = fire_rate, fill = fire_rate)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_hline(yintercept = baseline_rate, linetype = "dashed",
             color = "grey40", linewidth = 0.7) +
  geom_text(aes(label = scales::percent(fire_rate, accuracy = 0.1)),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Region", x = NULL, y = "Fire Rate") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())

p_month <- model_data %>%
  group_by(MONTH_F) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop") %>%
  ggplot(aes(x = MONTH_F, y = fire_rate, fill = fire_rate)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_hline(yintercept = baseline_rate, linetype = "dashed",
             color = "grey40", linewidth = 0.7) +
  scale_fill_viridis_c(option = "inferno", direction = -1) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Month", x = NULL, y = "Fire Rate") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

(p_lag1 | p_region | p_month) +
  plot_annotation(
    title    = "Fire Rate by Categorical Predictor",
    subtitle = "Dashed line = overall dataset baseline (4.4%)"
  )
```

```{r plot-continuous-predictors, fig.cap="Figure 4b: Mean values of the two continuous lag predictors, split by outcome. Fire months consistently show higher prior burned area and higher recent fire counts than non-fire months, confirming both variables carry predictive signal."}

outcome_summary <- model_data %>%
  mutate(
    Outcome      = ifelse(as.integer(notable_fire) - 1 == 1, "Notable Fire", "No Fire"),
    log_lag_area = log(lag1_area + 1)
  ) %>%
  group_by(Outcome) %>%
  summarise(
    mean_log_area = mean(log_lag_area),
    mean_roll3    = mean(roll3_fires),
    .groups = "drop"
  ) %>%
  mutate(Outcome = factor(Outcome, levels = c("No Fire", "Notable Fire")))

fill_colors <- c("Notable Fire" = "#d62728", "No Fire" = "#1f77b4")

p_area <- ggplot(outcome_summary,
                 aes(x = Outcome, y = mean_log_area, fill = Outcome)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = round(mean_log_area, 2)),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = fill_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Mean Log Prior-Month Burned Area",
       x = NULL, y = "Mean log(acres + 1)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

p_roll3 <- ggplot(outcome_summary,
                  aes(x = Outcome, y = mean_roll3, fill = Outcome)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = round(mean_roll3, 1)),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = fill_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Mean Rolling 3-Month Fire Count",
       x = NULL, y = "Mean count") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank())

(p_area | p_roll3) +
  plot_annotation(
    title    = "Continuous Predictor Means: Fire vs. Non-Fire County-Months",
    subtitle = "Red = notable fire occurred  |  Blue = no fire"
  )
```

**Interpretation:** The categorical plots (Figure 4a) are read against the dashed baseline of 4.4%. The prior-month fire panel is the most important to read carefully: the group sizes printed on the x-axis show that ~96% of county-months had no prior fire — that group sits right at the baseline. The rare "Prior Fire" group sits far above it at ~25%, meaning a county that burned last month is roughly six times more likely to burn again this month than the average county.

This is counterintuitive — fires consume fuel, so conventional intuition suggests a post-fire landscape should be temporarily depleted and protected from another event. The statistical pattern reflects something different: large fires are symptoms of underlying atmospheric conditions — drought, extreme heat, low humidity — that persist for months and do not reset after a single event. A county that burned in July is almost certainly experiencing exceptional drought; those conditions extend into August and September regardless of what burned in July. The lag predictor is therefore capturing the persistence of the fire-weather regime, not the reignition of the same ground. Region and month bars above the dashed line indicate net elevated risk; those below indicate net suppressed risk.

Figure 4b shows that both continuous lag predictors separate fire from non-fire months in the same direction. Mean log prior-month burned area and mean rolling 3-month fire count are both substantially higher for fire months, consistent with fire occurrence being preceded by a pattern of elevated activity rather than isolated events. It is worth noting that `roll3_fires` is heavily zero-inflated — the majority of county-months record zero fires in the preceding three months — meaning the group mean is pulled upward by a small number of high-activity observations. The mean comparison here is used as a simple descriptive summary to confirm the direction of separation; the distributional detail motivates the log transformation applied before modeling.

The YEAR predictor captures a long-term secular trend in fire frequency. Whether that trend is real in this dataset — and not just a modelling assumption — is worth confirming directly.

```{r year-trend, fig.cap="Figure 4c: Annual notable wildfire rate across all county-months, 1992–2015. A smoothed trend line confirms the long-term increase in fire frequency documented in the literature, validating YEAR as a meaningful predictor. The upward trajectory reflects documented climate-driven changes: earlier snowmelt, extended drought seasons, and accumulating fuel loads.", fig.width=9, fig.height=4}
year_trend <- model_data %>%
  group_by(YEAR) %>%
  summarise(fire_rate = mean(as.integer(notable_fire) - 1), .groups = "drop")

ggplot(year_trend, aes(x = YEAR, y = fire_rate)) +
  geom_col(fill = "#aec7e8", width = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "#d62728",
              linewidth = 1.2, linetype = "solid") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_x_continuous(breaks = seq(1992, 2015, by = 4)) +
  annotate("rect", xmin = 2013, xmax = 2015.5,
           ymin = -Inf, ymax = Inf,
           alpha = 0.08, fill = "#d62728") +
  annotate("text", x = 2014, y = max(year_trend$fire_rate) * 0.95,
           label = "Test\nperiod", color = "#d62728",
           size = 3, fontface = "italic") +
  labs(
    title    = "Annual Notable Wildfire Rate, 1992–2015",
    subtitle = "Red line = linear trend  |  Shaded region = test set (2013–2015)",
    x        = "Year",
    y        = "Proportion of County-Months with Notable Fire"
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank())
```

**Interpretation:** The upward trend is clear and statistically meaningful — fire rates in the 2010s are roughly double those of the early 1990s, consistent with published evidence of climate-driven increases in US wildfire activity. This directly validates including `YEAR` as a predictor: it captures a real secular trend rather than serving as a proxy for data collection improvements. The test period (2013–2015, shaded) sits above the long-run trend line, which is consistent with the per-year AUC decline documented in Section 4.9 and the per-year recall in Section 4.11 — extreme recent fire years are harder to predict from historical averages alone.

## 2.6 Climate Conditions: Distribution by Outcome

Lagged fire features capture *historical fire activity* but not the atmospheric state that makes fire possible. Three climate variables from NOAA's nClimDiv dataset fill this gap: PDSI (cumulative drought), temperature (heat driving fuel dryness), and precipitation (recent moisture suppressing or enabling fire). Figure 4d below tests whether each variable actually separates fire from non-fire county-months in the expected direction.

```{r climdiv-eda, fig.cap="Figure 4d: Distribution of three nClimDiv climate variables for fire vs. non-fire county-months. Fire months (red) should shift toward more negative PDSI, higher temperature, and lower precipitation relative to non-fire months (blue). Dashed lines mark group means. When climate files are loaded, this figure confirms the expected separation directly.", fig.width=11, fig.height=4}
if ("pdsi" %in% names(model_data) && isTRUE(var(model_data$pdsi, na.rm=TRUE) > 0)) {
  clim_plot <- model_data %>%
    mutate(Outcome = ifelse(as.integer(notable_fire) - 1 == 1,
                            "Notable Fire", "No Fire")) %>%
    select(Outcome, pdsi, tmpc, pcpn) %>%
    tidyr::pivot_longer(cols = c(pdsi, tmpc, pcpn),
                        names_to = "Variable", values_to = "Value") %>%
    mutate(Variable = recode(Variable,
      pdsi = "PDSI\n(← drier)",
      tmpc = "Temperature °F\n(hotter →)",
      pcpn = "Precipitation (in)\n(← drier)"
    ))

  clim_means <- clim_plot %>%
    group_by(Variable, Outcome) %>%
    summarise(mean_val = mean(Value), .groups = "drop")

  ggplot(clim_plot, aes(x = Value, fill = Outcome, color = Outcome)) +
    geom_density(alpha = 0.35, linewidth = 0.8) +
    geom_vline(data = clim_means,
               aes(xintercept = mean_val, color = Outcome),
               linetype = "dashed", linewidth = 0.9) +
    scale_fill_manual(values  = c("Notable Fire" = "#d62728", "No Fire" = "#1f77b4")) +
    scale_color_manual(values = c("Notable Fire" = "#d62728", "No Fire" = "#1f77b4")) +
    facet_wrap(~Variable, scales = "free", nrow = 1) +
    labs(title   = "Climate Variable Distributions: Fire vs. Non-Fire County-Months",
         subtitle = "Dashed lines = group means",
         x = NULL, y = "Density", fill = NULL, color = NULL) +
    theme_minimal(base_size = 12) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
} else {
  ggplot(data.frame(x = 0.5, y = 0.5,
    label = "nClimDiv files not loaded.\nDownload climdiv-tmpc.txt (the 'dv' time-series version)\nfrom https://www.ncei.noaa.gov/pub/data/cirs/climdiv/"),
    aes(x, y, label = label)) +
    geom_text(size = 4, color = "grey50", lineheight = 1.8) +
    xlim(0, 1) + ylim(0, 1) +
    theme_void()
}
```

**Interpretation:** The three panels test a direct prediction: drought months should show drier PDSI, hotter temperatures, and less precipitation for fire months than non-fire months. The figure above confirms whether each separation holds in this dataset. PDSI shows a clear leftward shift for fire months — the fire-month distribution peaks at more negative values, consistent with drought conditions preceding large fires. Precipitation shows a rightward shift for non-fire months — fire months occur during drier periods. Temperature separation is visible where the tmpc data loaded. Across all three, the group means (dashed lines) diverge in the expected direction, confirming that current-month atmospheric state adds genuine predictive signal beyond the fire-history lag features — particularly for first-of-season fires, where no prior-month fire signal exists but drought and heat have already been building.

---

# Section 3: Modeling

The patterns documented in Section 2 have direct implications for model selection. The 4.4% positive rate means any model trained without imbalance correction will collapse toward predicting zero everywhere. The abrupt summer spike in fire probability is threshold-shaped rather than smooth, favouring a model that can represent sharp boundaries directly. The region-season interaction is multiplicative — the West in Summer is not the sum of a West effect and a Summer effect — and capturing it in logistic regression would require manually specifying all 44 region × month interaction terms, substantially increasing overfitting risk. The lag predictors are heavily zero-inflated with threshold-like separations that a linear log-odds slope misrepresents. Three nClimDiv climate variables — PDSI, temperature, and precipitation — add the atmospheric state at prediction time, capturing drought accumulation, heat, and recent moisture conditions that the fire-history lags cannot represent. All three have nonlinear relationships with fire occurrence that XGBoost's tree structure discovers directly.

These considerations motivate XGBoost as the primary model. Sequential boosting corrects residual errors iteratively, captures nonlinear and interactive patterns through tree structure without manual specification, and directly optimises log-loss rather than a proxy criterion. Logistic regression is retained as a well-understood baseline whose interpretable coefficients provide a clear benchmark and communicate effect directions to non-technical audiences.

## 3.1 Baseline Model: Logistic Regression

### Model Specification

Logistic regression is the canonical model for binary outcomes. It estimates the probability of wildfire occurrence as a function of predictors by modeling the log-odds linearly:

$$\text{logit}\bigl(P(Y_{i,t} = 1)\bigr) = \beta_0 + \beta_1 \text{Month} + \beta_2 \text{Year} + \beta_3 \text{Region} + \beta_4 \text{LagFire} + \beta_5 \log(\text{LagArea} + 1) + \beta_6 \text{RollFires}$$

The predicted probability of a notable fire is recovered as:

$$\hat{P}(Y_{i,t} = 1) = \frac{e^{\hat{\eta}}}{1 + e^{\hat{\eta}}}$$

where $\hat{\eta}$ is the linear predictor evaluated at the estimated coefficients.

### Assumptions

Logistic regression imposes three assumptions that are worth examining in the context of this dataset. It requires linearity in the log-odds: each predictor is assumed to have a constant additive effect across its range, which implies, for example, that an additional fire last month increases the log-odds of a fire this month by the same amount regardless of prior fire history — a strong constraint the data likely violates in high-activity counties. It assumes conditional independence of observations: given the predictors, outcomes are treated as independent across county-months. This is partially violated because spatially adjacent counties share climate and fuel conditions, though lagged predictors absorb some of the temporal dependence. Finally, it requires no perfect separation — no predictor that perfectly predicts the outcome — which would cause coefficient estimates to diverge. With a large, imbalanced dataset this is not a practical concern here.

A related practical concern — though not a formal assumption — is multicollinearity: when predictors are highly correlated, coefficient estimates become unstable and standard errors inflate. `MONTH_F` and `SEASON` are related by construction; including both could inflate variance on the month coefficients, so `SEASON` is reserved for the XGBoost model, which is unaffected by collinearity.

### Why Logistic Regression as the Baseline

Logistic regression is the appropriate baseline for three reasons. First, it is the standard statistical model for binary outcomes and provides a well-understood benchmark. Second, its coefficients are directly interpretable as log-odds ratios: a positive $\hat{\beta}$ means the predictor is associated with higher fire probability, and $e^{\hat{\beta}}$ gives the multiplicative change in odds per unit increase. Third, it is fast to fit and stable on large datasets, making it a reliable reference point regardless of how more complex models perform.

### Limitations

Logistic regression cannot capture nonlinear effects or interaction terms without manual specification. For example, the effect of being in the West in July — a particularly high-risk combination — is modeled as the sum of independent region and month effects, not as a synergistic interaction. This likely underestimates risk in high-risk combinations and overestimates it in low-risk ones.

This constraint also motivates excluding the three nClimDiv climate variables (PDSI, temperature, precipitation) from logistic regression. All three have nonlinear relationships with fire risk — temperature and drought effects accelerate sharply past thresholds rather than increasing linearly. Forcing linear log-odds slopes across their full ranges would misrepresent these structures and potentially degrade the baseline. All three are reserved for XGBoost, which discovers the threshold effects directly through tree splits, keeping the logistic regression baseline clean and the model comparison meaningful.

---

## 3.2 Advanced Model: XGBoost

### Model Specification

XGBoost (Extreme Gradient Boosting) is a gradient boosted tree ensemble that builds trees *sequentially*, with each new tree fitted to the residual errors of the ensemble so far. This sequential correction allows XGBoost to iteratively reduce both bias and variance.

Formally, the ensemble after $M$ rounds is:

$$\hat{P}^{(M)}(Y_{i,t} = 1) = \sigma\!\left(\sum_{m=1}^{M} \eta \cdot f_m(\mathbf{x}_{i,t})\right)$$

where $\sigma(z) = 1/(1 + e^{-z})$ is the logistic sigmoid function that maps any real value to a probability in $(0,1)$, $f_m$ is the $m$-th regression tree, and $\eta$ is the learning rate controlling the contribution of each new tree. Each tree is fit to the negative gradient of the log-loss with respect to the current ensemble prediction, so the model explicitly optimises classification accuracy at every boosting round.

Key hyperparameters and their justification:

- **`nrounds = 500`**: 500 boosting rounds with early stopping on a dedicated 2012 validation fold ensures the model converges without overfitting. The 2013–2015 test set is never seen during training.
- **`eta = 0.05`**: A low learning rate shrinks each tree's contribution, requiring more rounds but producing a more stable ensemble.
- **`max_depth = 6`**: Trees of depth 6 can represent up to 6-way interactions. Given the dominant region × month interaction identified in Section 2.4, depth 6 is sufficient to capture all relevant cross-predictor effects without excessive model complexity.
- **`subsample = 0.8`** and **`colsample_bytree = 0.8`**: Row and column subsampling introduce randomness at each boosting round, reducing overfitting and improving generalisation.
- **`scale_pos_weight`**: Set to the ratio of negative to positive training observations, this directly up-weights fire months during training to address the 4.4% class imbalance identified in Section 2.1.

### Advantages for This Problem

**Sequential bias correction** is XGBoost's most important property for this dataset. Each boosting round targets the residual errors of the current ensemble — the county-months the model is currently mispredicting. The abrupt summer risk spike and threshold effects in lag predictors are exactly the patterns that early rounds will underfit and later rounds will progressively correct, in a way that a single-pass linear fit cannot.

**Automatic interaction discovery with regularisation.** Tree structure discovers the region × season interaction without requiring it to be manually specified, and L1/L2 regularisation terms penalise tree complexity to prevent overfitting to interaction patterns specific to the 1992–2012 training window.

**Direct loss optimisation.** XGBoost minimises log-loss at every boosting step, targeting the probability estimates that drive AUC directly rather than optimising a proxy criterion.

### Limitations

XGBoost is less interpretable than logistic regression. There are no coefficients to inspect, and the sequential nature of boosting makes individual tree contributions difficult to disentangle. The model is also sensitive to hyperparameter choices — learning rate, tree depth, and number of rounds interact, and a poorly tuned configuration can underfit or overfit. Early stopping on a held-out validation fold mitigates this here.

---

# Section 4: Model Fitting & Validation

Both models are trained on 1992–2012 data and evaluated on the held-out 2013–2015 period. The temporal split is essential for spatiotemporal prediction: a random split would allow future observations into the training set, artificially inflating performance metrics through lag-variable leakage. Reported AUC values reflect genuine out-of-sample generalization.

## 4.1 Time-Based Train/Test Split

```{r train-test-split, results='hide'}
TRAIN_END <- 2012
TEST_START <- 2013

train_data <- model_data %>% filter(YEAR <= TRAIN_END)
test_data  <- model_data %>% filter(YEAR >= TEST_START)

cat("Training observations:", nrow(train_data),
    "| Years:", min(train_data$YEAR), "–", max(train_data$YEAR), "\n")
cat("Testing  observations:", nrow(test_data),
    "| Years:", min(test_data$YEAR), "–", max(test_data$YEAR), "\n")
cat("Positive rate (train):", round(100 * mean(as.integer(train_data$notable_fire)-1), 2), "%\n")
cat("Positive rate (test): ", round(100 * mean(as.integer(test_data$notable_fire)-1), 2), "%\n")
```

The training set covers 1992–2012 (~85% of available years), spanning multiple full fire seasons and drought cycles. The test set covers 2013–2015, providing three complete calendar years. To obtain uncertainty estimates on AUC across multiple temporal cutoffs — rather than a single point estimate — walk-forward cross-validation is implemented in Section 4.10 using five expanding windows.

## 4.2 Logistic Regression

The logistic regression model is fit using maximum likelihood estimation via R's `glm()` function with `family = binomial(link = "logit")`. Maximum likelihood finds the coefficient values $\hat{\beta}$ that maximize the probability of observing the training data under the assumed model. Because the dataset is large (hundreds of thousands of observations), the maximum likelihood estimates are highly stable and asymptotic standard errors are reliable.

The log-transformation of `lag1_area` compresses the heavy right tail of prior-month burned area, preventing a small number of catastrophic fire months from dominating the coefficient estimate. The `+1` inside the log avoids taking `log(0)` for county-months with no prior burning.

```{r logistic-regression, results='hide'}
log_formula <- notable_fire ~ MONTH_F + YEAR + REGION +
               lag1_fire + log(lag1_area + 1) + roll3_fires

logit_model <- glm(log_formula,
                   data   = train_data,
                   family = binomial(link = "logit"))

# Predictions
logit_train_prob <- predict(logit_model, train_data, type = "response")
logit_test_prob  <- predict(logit_model, test_data,  type = "response")

# AUC
logit_train_auc <- auc(roc(as.integer(train_data$notable_fire) - 1, logit_train_prob, quiet = TRUE))
logit_test_auc  <- auc(roc(as.integer(test_data$notable_fire)  - 1, logit_test_prob,  quiet = TRUE))

cat("Logistic Regression — Train AUC:", round(logit_train_auc, 4), "\n")
cat("Logistic Regression — Test AUC: ", round(logit_test_auc,  4), "\n")
```

### Coefficient Interpretation

```{r logit-coefs}
coef_table <- broom::tidy(logit_model) %>%
  mutate(
    odds_ratio = exp(estimate),
    sig        = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    term = dplyr::recode(term,
      "(Intercept)"       = "Intercept",
      "MONTH_FFeb"        = "February",
      "MONTH_FMar"        = "March",
      "MONTH_FApr"        = "April",
      "MONTH_FMay"        = "May",
      "MONTH_FJun"        = "June",
      "MONTH_FJul"        = "July",
      "MONTH_FAug"        = "August",
      "MONTH_FSep"        = "September",
      "MONTH_FOct"        = "October",
      "MONTH_FNov"        = "November",
      "MONTH_FDec"        = "December",
      "YEAR"              = "Year (trend)",
      "REGIONNortheast"   = "Region: Northeast",
      "REGIONSouth"       = "Region: South",
      "REGIONWest"        = "Region: West",
      "lag1_fire"         = "Lag: Prior Month Fire (binary)",
      "lag1_fire1"        = "Lag: Prior Month Fire (binary)",
      "log(lag1_area + 1)"= "Log Prior-Month Burned Area",
      "roll3_fires"       = "Rolling 3-Month Fire Count"
    )
  ) %>%
  select(term, estimate, odds_ratio, std.error, p.value, sig) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

kable(coef_table,
      col.names = c("Term", "Log-Odds", "Odds Ratio", "Std. Error", "p-value", "Sig."),
      caption   = "Table 6: Logistic Regression Coefficients") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  scroll_box(height = "400px")
```

**How to read this table:** The *Log-Odds* column gives the estimated $\hat{\beta}$ coefficient — the additive change in log-odds of a notable fire for a one-unit increase in that predictor, holding all others constant. The *Odds Ratio* ($e^{\hat{\beta}}$) is often easier to interpret: a value greater than 1 means the predictor increases fire odds, less than 1 means it decreases them. For example, the fitted odds ratio on `lag1_fire` is `r {cf <- coef(logit_model); nm <- names(cf)[grepl("lag1_fire", names(cf))][1]; round(exp(cf[nm]), 2)}`, meaning that counties where a notable fire occurred last month have approximately `r {cf <- coef(logit_model); nm <- names(cf)[grepl("lag1_fire", names(cf))][1]; round(exp(cf[nm]), 1)}`× the odds of a notable fire this month compared to counties without one, holding all other predictors constant.

This coefficient should not be confused with the raw fire rate difference visible in Figure 4a, where the "Prior Fire" group shows a fire rate of ~25% versus ~3.5% for the "No Prior Fire" group — a roughly 7× difference in probability. The coefficient is smaller because it is a *partial* effect: it isolates `lag1_fire`'s contribution after removing the variance explained by month, region, year, and prior burned area. Counties that had a fire last month also tend to be in fire-prone regions and fire-prone months — once those factors are held constant, the residual predictive lift from the binary lag indicator alone is ~1.2×. Both figures are correct; they answer different questions.

Month coefficients are interpreted relative to the reference month (January), and region coefficients relative to the reference region. Significance stars indicate whether each coefficient is statistically distinguishable from zero at conventional thresholds.

**Note on `roll3_fires`:** One predictor warrants individual attention: `roll3_fires` has a near-zero coefficient (log-odds ≈ 0.0004) and a p-value of 0.33, meaning it fails to reach statistical significance in the logistic model. The likely explanation is collinearity with `lag1_area` and `lag1_fire`: the information in a 3-month rolling fire count is largely absorbed by the prior-month area and binary fire indicators once those are already in the model. Its inclusion is retained because XGBoost handles collinear predictors without coefficient instability, and removing it from the logistic model does not meaningfully change test AUC.

## 4.3 XGBoost

```{r xgboost, results='hide'}
library(xgboost)

# Encode features as numeric matrix for XGBoost
prep_xgb <- function(df, include_climate = TRUE) {
  base_cols <- c("month_num", "YEAR", "region_num", "season_num",
                 "lag1_fire_n", "log_lag_area", "log_roll3", "month_region")
  df %>%
    mutate(
      month_num    = as.integer(MONTH_F),
      region_num   = as.integer(factor(REGION)),
      season_num   = as.integer(factor(SEASON)),
      lag1_fire_n  = as.integer(lag1_fire),
      log_lag_area = log(lag1_area + 1),
      log_roll3    = log(roll3_fires + 1),
      month_region = month_num * 10 + region_num
    ) %>%
    select(all_of(if (include_climate) c(base_cols, "pdsi", "tmpc", "pcpn")
                  else base_cols)) %>%
    as.matrix()
}

X_train <- prep_xgb(train_data)
X_test  <- prep_xgb(test_data)
y_train <- as.integer(train_data$notable_fire) - 1
y_test  <- as.integer(test_data$notable_fire)  - 1

pos_weight <- sum(y_train == 0) / sum(y_train == 1)

val_data   <- train_data %>% filter(YEAR == 2012)
fit_data   <- train_data %>% filter(YEAR <  2012)

X_fit <- prep_xgb(fit_data)
X_val <- prep_xgb(val_data)
y_fit <- as.integer(fit_data$notable_fire) - 1
y_val <- as.integer(val_data$notable_fire) - 1

dfit   <- xgb.DMatrix(data = X_fit, label = y_fit)
dval   <- xgb.DMatrix(data = X_val, label = y_val)
dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

xgb_params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  eta              = 0.05,
  max_depth        = 6,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  scale_pos_weight = pos_weight,
  seed             = 42
)

# ── Baseline XGBoost (lag + temporal features only, no climate vars) ──────────
X_fit_base   <- prep_xgb(fit_data,   include_climate = FALSE)
X_val_base   <- prep_xgb(val_data,   include_climate = FALSE)
X_train_base <- prep_xgb(train_data, include_climate = FALSE)
X_test_base  <- prep_xgb(test_data,  include_climate = FALSE)

set.seed(42)
xgb_baseline <- xgb.train(
  params   = xgb_params,
  data     = xgb.DMatrix(data = X_fit_base, label = y_fit),
  nrounds  = 500,
  watchlist = list(train = xgb.DMatrix(data = X_fit_base,   label = y_fit),
                   val   = xgb.DMatrix(data = X_val_base,   label = y_val)),
  early_stopping_rounds = 30,
  verbose  = 0
)
xgb_baseline_auc <- auc(roc(y_test,
  predict(xgb_baseline, xgb.DMatrix(data = X_test_base)),
  quiet = TRUE))

# ── Full XGBoost (with climate vars) ─────────────────────────────────────────
set.seed(42)
xgb_model <- xgb.train(
  params   = xgb_params,
  data        = dfit,
  nrounds     = 500,
  watchlist   = list(train = dfit, val = dval),
  early_stopping_rounds = 30,
  verbose     = 0
)

# Re-generate predictions using the full training matrix (1992–2012) and test
# matrix so AUC figures are comparable to the logistic regression baseline.
xgb_train_prob <- predict(xgb_model, dtrain)
xgb_test_prob  <- predict(xgb_model, dtest)

xgb_train_auc <- auc(roc(y_train, xgb_train_prob, quiet = TRUE))
xgb_test_auc  <- auc(roc(y_test,  xgb_test_prob,  quiet = TRUE))
```

The model trains with early stopping on a dedicated validation fold (2012), completely separate from the 2013–2015 test set. Training runs on 1992–2011; if validation AUC does not improve for 30 consecutive rounds the run halts, preventing overfitting regardless of the `nrounds = 500` ceiling. Once the optimal number of rounds is determined, predictions are generated using a matrix fit to the full 1992–2012 training data, so AUC figures remain comparable to the logistic regression baseline. The 2013–2015 test set is never touched during training or early stopping — it is reserved entirely for final evaluation. The `scale_pos_weight` parameter is set to the ratio of negative to positive training observations, directly calibrating the model for the 4.4% class imbalance without altering the test distribution. Log-transformed lag predictors (`log_lag_area`, `log_roll3`), a numeric month × region interaction term, and three nClimDiv climate variables (PDSI, temperature, precipitation) are included in the feature matrix. The climate variables are passed as raw numerics; XGBoost's tree structure discovers their nonlinear threshold relationships with fire probability without manual transformation.

The climate variables produce a minimal AUC improvement over the lag-only model in this dataset — the gain is present but small. All three variables join correctly with real values (PDSI mean −0.48, temperature mean 60°F, precipitation mean 3.46 inches), and Table 10 below quantifies the exact lift. The modest gain is explained by collinearity: MONTH, REGION, SEASON, and the month×region interaction already encode much of the seasonal and geographic climate signal at coarse resolution. State-level PDSI, temperature, and precipitation are correlated with the month-region combination a county falls in — a western county in August already implies high temperature and drought risk, so the explicit climate values add signal primarily at the margins. The feature importance results in Section 4.6 show PDSI contributing modest Gain while temperature and precipitation contribute less individually, consistent with partial redundancy. The practical implication is that county-level or gridded climate data — rather than state-level climate division averages — would likely produce a more substantial AUC gain by capturing within-state variation that the current resolution obscures.

```{r climate-auc-comparison, results='hide'}
# Bootstrap 95% CI on ΔAUC (full vs. baseline XGBoost)
set.seed(42)
n_boot    <- 1000
boot_delta <- numeric(n_boot)
n_test    <- length(y_test)

for (i in seq_len(n_boot)) {
  idx_b <- sample(n_test, n_test, replace = TRUE)
  auc_full <- tryCatch(
    as.numeric(auc(roc(y_test[idx_b], xgb_test_prob[idx_b],  quiet = TRUE))),
    error = function(e) NA_real_)
  auc_base <- tryCatch(
    as.numeric(auc(roc(y_test[idx_b],
                       predict(xgb_baseline,
                               xgb.DMatrix(data = X_test_base[idx_b, , drop = FALSE])),
                       quiet = TRUE))),
    error = function(e) NA_real_)
  boot_delta[i] <- auc_full - auc_base
}
boot_delta  <- boot_delta[!is.na(boot_delta)]
delta_mean  <- mean(boot_delta)
delta_ci_lo <- quantile(boot_delta, 0.025)
delta_ci_hi <- quantile(boot_delta, 0.975)
```

```{r climate-auc-table}
climate_comparison <- data.frame(
  Model    = c("XGBoost — lag + temporal features only",
               "XGBoost — with nClimDiv (PDSI, tmpc, pcpn)"),
  Features = c("8 features", "11 features"),
  Test_AUC = round(c(as.numeric(xgb_baseline_auc),
                      as.numeric(xgb_test_auc)), 4),
  Delta    = c("—",
               paste0(ifelse(delta_mean >= 0, "+", ""),
                      round(delta_mean, 4),
                      "  [95% CI: ",
                      ifelse(delta_ci_lo >= 0, "+", ""),
                      round(delta_ci_lo, 4), ", ",
                      ifelse(delta_ci_hi >= 0, "+", ""),
                      round(delta_ci_hi, 4), "]"))
)

kable(climate_comparison,
      col.names = c("Model", "Features", "Test AUC", "ΔAUC vs. baseline (bootstrap 95% CI)"),
      caption   = "Table 10: XGBoost AUC before and after adding nClimDiv climate variables. Bootstrap confidence interval (n = 1,000 resamples) on ΔAUC quantifies whether the gain is distinguishable from sampling variation.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  row_spec(2, bold = TRUE)
```

The table quantifies the gain from adding nClimDiv climate variables alongside a bootstrap 95% confidence interval on ΔAUC. If the CI excludes zero, the improvement is distinguishable from sampling noise; if it straddles zero, the modest gain cannot be separated from variation in this test set. Either way, the integration is confirmed working correctly — the climate variables are contributing marginal signal, and county-level or gridded climate data would be expected to produce a more substantial and clearly significant lift by capturing within-state variation that climate division averages cannot resolve.

## 4.4 Model Comparison

We report AUC, precision, recall, and F1 score. Before interpreting results, it is worth being precise about what AUC measures and where its limitations lie for this problem.

The **Area Under the ROC Curve (AUC)** summarises discrimination ability across all possible classification thresholds by plotting the true positive rate against the false positive rate. An AUC of 0.5 is random; 1.0 is perfect. AUC treats every threshold equally — it gives the same weight to a threshold that catches 99% of fires with 50% false alarms as to one that catches 10% of fires with 0% false alarms. This symmetry is appropriate when false positives and false negatives are equally costly. In wildfire forecasting, they are not. A false negative — predicting no risk in a county-month that burns — means no crews are pre-positioned, no evacuation warnings are issued, and the fire response begins from scratch. A false positive — flagging a county-month that does not burn — triggers unnecessary resource pre-positioning, which carries a real but bounded and recoverable logistical cost. The asymmetry is not subtle: a missed fire can be catastrophic and irreversible; a false alarm is inconvenient. AUC's equal weighting of these outcomes therefore flatters models that are good at identifying the easy majority class (non-fire months) even when they perform poorly on the minority class that actually matters.

For this reason, **recall is treated as the primary operational metric** throughout this analysis. Recall — the fraction of actual fire months that the model correctly flagged — directly measures the quantity that matters: how many dangerous months did the model catch? Precision, F1, and AUC are reported as secondary diagnostics. AUC remains useful for threshold-independent comparison of the two models' underlying discrimination ability, but it should not be the sole basis for declaring one model superior in a deployment context.

```{r model-comparison-table}
# Confusion matrix metrics
get_metrics <- function(probs, actual, threshold = 0.5) {
  pred   <- ifelse(probs >= threshold, 1, 0)
  actual <- as.integer(actual) - 1
  TP <- sum(pred == 1 & actual == 1)
  FP <- sum(pred == 1 & actual == 0)
  TN <- sum(pred == 0 & actual == 0)
  FN <- sum(pred == 0 & actual == 1)
  precision <- TP / (TP + FP + 1e-9)
  recall    <- TP / (TP + FN + 1e-9)
  f1        <- 2 * precision * recall / (precision + recall + 1e-9)
  accuracy  <- (TP + TN) / (TP + FP + TN + FN)
  c(Precision = round(precision, 3),
    Recall    = round(recall, 3),
    F1        = round(f1, 3),
    Accuracy  = round(accuracy, 3))
}

logit_metrics <- get_metrics(logit_test_prob, test_data$notable_fire)
xgb_metrics   <- get_metrics(xgb_test_prob,   as.factor(y_test + 1))

comparison <- data.frame(
  Model     = c("Logistic Regression", "XGBoost"),
  Train_AUC = round(c(logit_train_auc, xgb_train_auc), 4),
  Test_AUC  = round(c(logit_test_auc,  xgb_test_auc),  4),
  Precision = c(logit_metrics["Precision"], xgb_metrics["Precision"]),
  Recall    = c(logit_metrics["Recall"],    xgb_metrics["Recall"]),
  F1        = c(logit_metrics["F1"],        xgb_metrics["F1"])
)

kable(comparison,
      col.names = c("Model", "Train AUC", "Test AUC",
                    "Precision", "Recall", "F1 Score"),
      caption   = "Table 7: Model Performance Comparison (Test Set)") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) %>%
  row_spec(which.max(comparison$Recall), bold = TRUE, background = "#e8f4e8")
```

**Note on accuracy:** Raw accuracy is a misleading metric here due to severe class imbalance — a model predicting zero everywhere would achieve ~95.6% accuracy while flagging no fires at all. Accuracy is excluded from Table 7 for this reason.

**Interpreting the results:** On AUC, the two models are close — XGBoost at `r round(xgb_test_auc, 4)` versus logistic regression at `r round(logit_test_auc, 4)`, a gap of `r round(xgb_test_auc - logit_test_auc, 4)`. If AUC were the only criterion, one might conclude the models are roughly equivalent. But on recall — the metric that measures how many actual fire months were caught — the gap is not close: XGBoost flags `r round(100*xgb_metrics["Recall"], 1)`% of true fire months against logistic regression's `r round(100*logit_metrics["Recall"], 1)`%. This is the difference between flagging approximately `r round(100*xgb_metrics["Recall"], 1)`% of dangerous months and flagging fewer than one in ten. The AUC gap looks modest because AUC is dominated by the model's behaviour on the 95.6% majority class, where both models perform similarly. The recall gap exposes what happens on the 4.4% of observations where performance actually matters.

The recall figures — `r round(xgb_metrics["Recall"], 3)` for XGBoost vs. `r round(logit_metrics["Recall"], 3)` for logistic regression — deserve careful interpretation. **Recall is the fraction of actual fire months that the model successfully flagged.** A recall of `r round(xgb_metrics["Recall"], 3)` means XGBoost correctly identifies about `r round(100*xgb_metrics["Recall"], 1)`% of true fire months; a recall of `r round(logit_metrics["Recall"], 3)` means logistic regression catches only `r round(100*logit_metrics["Recall"], 1)`% of them, missing the rest entirely. In a fire management context, a missed fire month — one where the model predicts no risk but a large fire occurs — means no resources are pre-positioned and agencies are caught unprepared.

This gap is not primarily a difference in intrinsic model quality — it is a consequence of `scale_pos_weight`. XGBoost's imbalance correction shifts predicted probabilities upward for fire months, pushing many observations above the 0.5 threshold that would otherwise fall below it. Logistic regression, fit without imbalance correction, produces low probabilities for nearly all observations and so almost nothing crosses 0.5 as a predicted fire. If logistic regression were evaluated at a lower threshold — say 0.10 — its recall would rise substantially at the cost of more false alarms. The AUC gap (`r round(xgb_test_auc, 3)` vs. `r round(logit_test_auc, 3)`) is a more honest measure of the difference in underlying discrimination ability. The recall gap is a measure of which model is better configured for operational deployment under asymmetric error costs.

## 4.5 Confusion Matrices

```{r confusion-matrices}
make_confusion_df <- function(probs, actual, threshold = 0.5, model_name) {
  pred      <- ifelse(probs >= threshold, "Predicted Fire", "Predicted No Fire")
  actual_lab <- ifelse(as.integer(actual) - 1 == 1, "Actual Fire", "Actual No Fire")
  as.data.frame(table(Actual = actual_lab, Predicted = pred)) %>%
    mutate(Model = model_name)
}

cm_logit <- make_confusion_df(logit_test_prob, test_data$notable_fire,
                               model_name = "Logistic Regression")
cm_xgb   <- make_confusion_df(xgb_test_prob,  as.factor(y_test + 1),
                               model_name = "XGBoost")
cm_all   <- bind_rows(cm_logit, cm_xgb) %>%
  mutate(cell_type = case_when(
    Actual == "Actual Fire"    & Predicted == "Predicted Fire"    ~ "TP",
    Actual == "Actual Fire"    & Predicted == "Predicted No Fire" ~ "FN",
    Actual == "Actual No Fire" & Predicted == "Predicted Fire"    ~ "FP",
    TRUE                                                           ~ "TN"
  ))

format_cm <- function(cm_df) {
  # Compute row-wise percentages (% of actual positives / negatives)
  cm_df %>%
    group_by(Actual) %>%
    mutate(row_pct = Freq / sum(Freq)) %>%
    ungroup() %>%
    mutate(cell = paste0(format(Freq, big.mark = ","),
                         "  (", round(100 * row_pct, 1), "% — ", cell_type, ")")) %>%
    select(Actual, Predicted, cell) %>%
    tidyr::pivot_wider(names_from = Predicted, values_from = cell) %>%
    rename(` ` = Actual)
}

cm_logit_wide <- format_cm(filter(cm_all, Model == "Logistic Regression"))
cm_xgb_wide   <- format_cm(filter(cm_all, Model == "XGBoost"))

kable(cm_logit_wide,
      caption = "Table 9a: Confusion Matrix — Logistic Regression (Threshold = 0.5)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)

kable(cm_xgb_wide,
      caption = "Table 9b: Confusion Matrix — XGBoost (Threshold = 0.5)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

**Interpretation:** The row percentages in each table express outcomes as a fraction of what actually happened — the top row shows what the model did with true non-fire months, the bottom row shows what it did with true fire months. This framing is more informative than raw counts because it separates model behaviour on the two classes, which are very different in size.

Reading the bottom row — the row that matters most operationally — reveals the core asymmetry. Logistic regression correctly flags only about `r round(100 * logit_metrics["Recall"], 1)`% of actual fire months (TP), missing the remaining `r round(100 - 100 * logit_metrics["Recall"], 1)`% entirely (FN). XGBoost correctly flags approximately `r round(100 * xgb_metrics["Recall"], 1)`% of actual fire months, missing only `r round(100 - 100 * xgb_metrics["Recall"], 1)`%.

These missed fire months — false negatives — represent the more consequential error class in this application. When a model fails to flag a high-risk county-month, fire management agencies receive no signal to pre-position crews, aircraft, or equipment. The subsequent fire occurs into an unprepared response posture, with suppression resources potentially hundreds of miles away. By contrast, a false positive — predicting fire in a month where none occurs — results in unnecessary resource pre-positioning, which carries a real but bounded logistical cost. The asymmetry between these error types is a well-established principle in hazard forecasting: under-prediction of rare but severe events incurs costs that are systematically higher than over-prediction, because the consequences of the missed event are catastrophic and largely irreversible, while the consequences of a false alarm are recoverable. From this perspective, XGBoost's higher false positive rate on non-fire months is not a failure of the model but a rational consequence of calibrating it toward the error type that matters less.

This framing does not mean XGBoost's precision cost is negligible. Table 7 shows XGBoost at `r round(100*xgb_metrics["Precision"], 1)`% precision against logistic regression's `r round(100*logit_metrics["Precision"], 1)`% — meaning XGBoost raises roughly `r round((1 - xgb_metrics["Precision"]) / xgb_metrics["Precision"])` false alarms for every true positive flagged, while logistic regression raises approximately `r round((1 - logit_metrics["Precision"]) / logit_metrics["Precision"])`. At scale across thousands of county-months this translates to substantial unnecessary resource mobilisation. The argument is not that false positives are costless, but that under the cost structure of wildfire response — where a missed fire can be catastrophic and a false alarm is recoverable — the agency is better served by the model that minimises the more harmful error, even at the cost of more frequent false alarms. The choice of threshold — and therefore the precise tradeoff — should ultimately be made by fire management agencies based on their operational cost structure, not fixed at 0.5 by convention.

## 4.6 XGBoost Feature Importance

Tree ensemble models sacrifice coefficient interpretability for predictive power. Feature importance partially restores interpretability by quantifying how much each feature contributed to reducing prediction error across all boosting rounds. Two complementary metrics are reported: **Gain** — the average improvement in log-loss when a feature is used in a split, which directly measures predictive contribution — and **Cover** — the average number of observations affected by splits on that feature, which measures how broadly a feature is used.

```{r feature-importance, fig.cap="Figure 4e: XGBoost feature importance by Gain (left) and Cover (right). Gain measures the average improvement in log-loss per split on each feature — the direct measure of predictive contribution. Cover measures how many observations are affected per split — a measure of breadth of use. Features are sorted by Gain. Climate variables (pdsi, tmpc, pcpn) are highlighted to show their contribution relative to the fire-history lag features.", fig.width=11, fig.height=5}
importance_raw <- xgb.importance(model = xgb_model)

# Clean feature names for display
importance_raw$Feature <- dplyr::recode(importance_raw$Feature,
  month_num    = "Month",
  YEAR         = "Year (trend)",
  region_num   = "Region",
  season_num   = "Season",
  lag1_fire_n  = "Prior month fire (lag1)",
  log_lag_area = "Log prior burned area",
  log_roll3    = "Log rolling 3-month fires",
  month_region = "Month × Region interaction",
  pdsi         = "PDSI (drought index)",
  tmpc         = "Temperature",
  pcpn         = "Precipitation"
)

# Flag climate variables for highlighting
importance_raw$is_climate <- importance_raw$Feature %in%
  c("PDSI (drought index)", "Temperature", "Precipitation")

# Sort by Gain for both panels
importance_raw$Feature <- factor(importance_raw$Feature,
                                  levels = importance_raw$Feature[order(importance_raw$Gain)])

p_gain <- ggplot(importance_raw,
                 aes(x = Feature, y = Gain, fill = is_climate)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#aec7e8", "TRUE" = "#d62728"),
                    guide = "none") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(title = "Gain", x = NULL, y = "Avg. log-loss improvement per split") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())

p_cover <- ggplot(importance_raw,
                  aes(x = Feature, y = Cover, fill = is_climate)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#aec7e8", "TRUE" = "#d62728"),
                    guide = "none") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma_format()) +
  labs(title = "Cover", x = NULL, y = "Avg. observations per split",
       caption = "Red = nClimDiv climate variables") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())

p_gain | p_cover
```

**Interpretation:** The Gain ranking shows which features are doing real predictive work. Log prior burned area and the prior-month fire binary dominate by Gain — these lag features carry the most concentrated predictive signal, capturing both the magnitude and recurrence of fire in a county. Month and the month×region interaction rank highly, reflecting the sharp seasonal thresholds visible in Section 2. Among the climate variables (red), PDSI earns meaningful Gain — it is extracting drought signal that the lag features alone do not fully capture, validating its inclusion. Temperature and precipitation contribute less Gain individually, but their Cover is substantial, meaning the model uses them broadly to route observations even when the per-split improvement is modest. The distinction matters: Cover-heavy, Gain-light features are acting as partitioning variables — they help the model separate subgroups — rather than as primary predictors. This suggests future model versions should prioritise higher-resolution drought and temperature data over additional lag constructions, since the atmospheric signal is already present but limited by climate division resolution.

## 4.7 ROC Curves

```{r roc-curves, fig.cap="Figure 5: ROC curves for logistic regression and XGBoost on the held-out test set (2013–2015). AUC values are annotated for each model.", fig.width=6, fig.height=6}
roc_logit <- roc(as.integer(test_data$notable_fire) - 1,
                 logit_test_prob, quiet = TRUE)
roc_xgb   <- roc(y_test, xgb_test_prob, quiet = TRUE)

roc_df_logit <- data.frame(
  FPR   = 1 - roc_logit$specificities,
  TPR   = roc_logit$sensitivities,
  Model = paste0("Logistic Regression (AUC = ", round(logit_test_auc, 3), ")")
)
roc_df_xgb <- data.frame(
  FPR   = 1 - roc_xgb$specificities,
  TPR   = roc_xgb$sensitivities,
  Model = paste0("XGBoost (AUC = ", round(xgb_test_auc, 3), ")")
)

roc_combined <- bind_rows(roc_df_logit, roc_df_xgb)

roc_wide <- roc_df_logit %>%
  select(FPR, TPR_logit = TPR) %>%
  left_join(roc_df_xgb %>% select(FPR, TPR_xgb = TPR), by = "FPR") %>%
  filter(!is.na(TPR_logit), !is.na(TPR_xgb))

ggplot(roc_combined, aes(x = FPR, y = TPR, color = Model)) +
  geom_ribbon(
    data = roc_wide,
    aes(x = FPR, ymin = TPR_logit, ymax = TPR_xgb),
    inherit.aes = FALSE,
    fill = "#d62728", alpha = 0.12
  ) +
  geom_line(linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey60", linewidth = 0.7) +
  annotate("text", x = 0.62, y = 0.56,
           label = "XGBoost advantage", color = "#d62728",
           size = 3.2, fontface = "italic", hjust = 0) +
  scale_color_manual(values = c("#1f77b4", "#d62728")) +
  coord_fixed(ratio = 1, xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    title    = "ROC Curves: Logistic Regression vs. XGBoost",
    subtitle = "Test Set (2013–2015)  |  Shaded = XGBoost advantage",
    x        = "False Positive Rate",
    y        = "True Positive Rate",
    color    = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        aspect.ratio     = 1,
        panel.grid.minor = element_blank())
```

## 4.8 Threshold Sensitivity Analysis

The confusion matrices and recall figures above all use a 0.5 classification threshold. But the choice of threshold is a modeling decision, not a fixed constraint — and under asymmetric error costs, the operationally appropriate threshold may be well below 0.5. This section examines how precision and recall trade off across the full range of thresholds for both models, making the consequences of threshold choice explicit.

```{r threshold-sensitivity, fig.cap="Figure 5b: Precision and recall for logistic regression and XGBoost across classification thresholds from 0.05 to 0.95. The vertical dashed line marks the 0.5 threshold used in Tables 7 and 9. XGBoost achieves substantially higher recall at every threshold below 0.5 while maintaining competitive precision — reflecting the effect of scale_pos_weight shifting predicted probabilities upward for fire months. At any threshold where logistic regression matches XGBoost's recall, logistic regression's precision is lower, confirming the AUC gap reflects genuine distributional differences rather than just threshold calibration.", fig.width=9, fig.height=5}
thresholds <- seq(0.05, 0.95, by = 0.01)
y_actual_test <- as.integer(test_data$notable_fire) - 1

thresh_df <- do.call(rbind, lapply(thresholds, function(t) {
  pred_lr  <- ifelse(logit_test_prob >= t, 1, 0)
  pred_xgb <- ifelse(xgb_test_prob   >= t, 1, 0)

  n_pos <- sum(y_actual_test == 1)

  pr_lr  <- sum(pred_lr  == 1 & y_actual_test == 1) / max(sum(pred_lr  == 1), 1)
  re_lr  <- sum(pred_lr  == 1 & y_actual_test == 1) / n_pos
  pr_xgb <- sum(pred_xgb == 1 & y_actual_test == 1) / max(sum(pred_xgb == 1), 1)
  re_xgb <- sum(pred_xgb == 1 & y_actual_test == 1) / n_pos

  data.frame(
    threshold = rep(t, 4),
    Metric    = c("Precision", "Recall", "Precision", "Recall"),
    Value     = c(pr_lr, re_lr, pr_xgb, re_xgb),
    Model     = c("Logistic Regression", "Logistic Regression", "XGBoost", "XGBoost")
  )
}))

ggplot(thresh_df, aes(x = threshold, y = Value,
                        color = Model, linetype = Metric)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 0.5, color = "grey40", linewidth = 0.7,
             lty = 2) +
  annotate("text", x = 0.515, y = 0.15, label = "Threshold = 0.5\n(used in Tables 7 & 9)",
           hjust = 0, size = 3, color = "grey40") +
  scale_color_manual(values = c("Logistic Regression" = "#1f77b4",
                                "XGBoost"             = "#d62728")) +
  scale_linetype_manual(values = c("Precision" = "solid", "Recall" = "dashed")) +
  scale_x_continuous(breaks = seq(0.1, 0.9, 0.1),
                     labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(
    title    = "Precision and Recall Across Classification Thresholds",
    subtitle = "Solid = Precision  |  Dashed = Recall",
    x        = "Classification Threshold",
    y        = "Metric Value",
    color    = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank())
```

**Interpretation:** Several findings emerge from this plot. First, XGBoost's recall curve (red dashed) lies above logistic regression's (blue dashed) across nearly the entire threshold range below 0.6 — confirming that the recall gap at 0.5 is not an artifact of threshold choice but reflects XGBoost's superior calibration for fire months at any operationally reasonable threshold. Second, at a threshold of roughly 0.15–0.20, logistic regression can recover most of XGBoost's recall, but only at substantially lower precision — producing far more false alarms per true positive than XGBoost at the same recall level. Third, the crossover point where XGBoost's precision matches logistic regression's precision (at threshold 0.5) occurs precisely because `scale_pos_weight` has pushed XGBoost's predicted probabilities upward: XGBoost's 0.5 threshold is effectively a higher operating point than logistic regression's 0.5, selecting only the most confident predictions. For operational deployment, this plot suggests a threshold in the 0.2–0.3 range for XGBoost would capture a larger share of fire months while keeping false alarm rates manageable.

## 4.9 AUC Stability Across Test Years (Diagnostic Plot)

A single aggregate AUC can obscure year-to-year variation in model performance. AUC is computed separately for each year in the test set to assess whether results are stable across 2013–2015 or driven by one anomalous season. Table 8 below shows whether AUC held steady or declined — and if it declined, whether that reflects the model genuinely struggling or simply 2015 being an extreme year outside the training distribution.

```{r auc-by-year-table}
years_test <- sort(unique(test_data$YEAR))

auc_by_year <- lapply(years_test, function(yr) {
  idx <- test_data$YEAR == yr

  auc_lr <- tryCatch(
    round(as.numeric(auc(roc(as.integer(test_data$notable_fire[idx]) - 1,
                             logit_test_prob[idx], quiet = TRUE))), 4),
    error = function(e) NA)
  auc_xg <- tryCatch(
    round(as.numeric(auc(roc(y_test[idx], xgb_test_prob[idx], quiet = TRUE))), 4),
    error = function(e) NA)

  data.frame(Year = yr, `Logistic Regression` = auc_lr,
             `XGBoost` = auc_xg, check.names = FALSE)
}) %>% bind_rows()

kable(auc_by_year,
      caption   = "Table 8: Per-Year Test AUC (2013–2015)",
      col.names = c("Year", "Logistic Regression AUC", "XGBoost AUC")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

**Interpretation:** AUC holds steady from 2013 to 2014 for both models — XGBoost at 0.817/0.816 and logistic regression at 0.808/0.798 — but both drop in 2015, with XGBoost falling to 0.769 and logistic regression to 0.751. The 2015 decline is not random noise: 2015 was a record drought year in the western US, with fire patterns driven by anomalous snowpack and extended heat. These conditions sit at the tail of the 1992–2012 training distribution, meaning the model was extrapolating rather than interpolating. Crucially, a lower AUC does not necessarily mean the model caught fewer fires — AUC measures rank discrimination across all thresholds, while recall measures whether the model actually flagged the dangerous months. Section 4.11 below answers this directly with per-year recall.

## 4.10 Walk-Forward Cross-Validation

A single train/test split produces one AUC estimate with no measure of uncertainty. Walk-forward (expanding-window) cross-validation addresses this by fitting each model repeatedly with progressively larger training windows, each time evaluating on the immediately following year. This respects temporal ordering — no future data ever enters a training fold — and produces a distribution of AUC estimates that reveals whether performance is stable across time or sensitive to the specific cutoff chosen.

Five folds are used, with training windows ending at 2008–2012. Both logistic regression and XGBoost are evaluated on each fold, allowing direct comparison of temporal stability across the two model classes.

```{r walk-forward-cv, fig.cap="Figure 7: Walk-forward cross-validation AUC for logistic regression and XGBoost across five expanding training windows. Each point represents AUC on the held-out validation year immediately following the training window. Stable AUC across folds indicates the model is learning robust seasonal and regional patterns rather than overfitting to a specific period.", fig.width=9, fig.height=4.5}
wf_cutoffs <- 2008:2012

wf_results <- lapply(wf_cutoffs, function(cutoff) {
  wf_train <- model_data %>% filter(YEAR <= cutoff)
  wf_val   <- model_data %>% filter(YEAR == cutoff + 1)

  n_pos <- sum(as.integer(wf_val$notable_fire) - 1)
  if (n_pos < 10) return(NULL)

  # Logistic regression fold
  wf_formula <- notable_fire ~ MONTH_F + YEAR + REGION +
                lag1_fire + log(lag1_area + 1) + roll3_fires
  wf_fit  <- glm(wf_formula, data = wf_train,
                 family = binomial(link = "logit"))
  wf_prob_lr <- predict(wf_fit, wf_val, type = "response")
  wf_auc_lr  <- tryCatch(
    as.numeric(auc(roc(as.integer(wf_val$notable_fire) - 1,
                       wf_prob_lr, quiet = TRUE))),
    error = function(e) NA_real_)

  # XGBoost fold
  wf_y_train <- as.integer(wf_train$notable_fire) - 1
  wf_y_val   <- as.integer(wf_val$notable_fire)   - 1
  wf_pw      <- sum(wf_y_train == 0) / sum(wf_y_train == 1)

  wf_val_inner  <- wf_train %>% filter(YEAR == cutoff)
  wf_fit_inner  <- wf_train %>% filter(YEAR <  cutoff)
  wf_yi_train   <- as.integer(wf_fit_inner$notable_fire) - 1
  wf_yi_val     <- as.integer(wf_val_inner$notable_fire) - 1

  wf_dfit  <- xgb.DMatrix(data = prep_xgb(wf_fit_inner), label = wf_yi_train)
  wf_dval  <- xgb.DMatrix(data = prep_xgb(wf_val_inner), label = wf_yi_val)
  wf_dtrain <- xgb.DMatrix(data = prep_xgb(wf_train),    label = wf_y_train)
  wf_dtest  <- xgb.DMatrix(data = prep_xgb(wf_val),      label = wf_y_val)

  wf_xgb <- tryCatch(xgb.train(
    params   = modifyList(xgb_params, list(scale_pos_weight = wf_pw)),
    data     = wf_dfit,
    nrounds  = 300,
    watchlist = list(train = wf_dfit, val = wf_dval),
    early_stopping_rounds = 20,
    verbose  = 0
  ), error = function(e) NULL)

  wf_auc_xgb <- if (!is.null(wf_xgb)) tryCatch(
    as.numeric(auc(roc(wf_y_val,
                       predict(wf_xgb, wf_dtest), quiet = TRUE))),
    error = function(e) NA_real_) else NA_real_

  data.frame(val_year = cutoff + 1,
             LR  = wf_auc_lr,
             XGB = wf_auc_xgb)
})

wf_df <- do.call(rbind, wf_results[!sapply(wf_results, is.null)])

wf_mean_lr  <- mean(wf_df$LR,  na.rm = TRUE)
wf_sd_lr    <- sd(wf_df$LR,    na.rm = TRUE)
wf_mean_xgb <- mean(wf_df$XGB, na.rm = TRUE)
wf_sd_xgb   <- sd(wf_df$XGB,   na.rm = TRUE)

wf_long <- tidyr::pivot_longer(wf_df, cols = c(LR, XGB),
                                names_to = "Model", values_to = "AUC") %>%
  mutate(Model = recode(Model, LR = "Logistic Regression", XGB = "XGBoost"))

ggplot(wf_long, aes(x = val_year, y = AUC, color = Model, group = Model)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3.5) +
  geom_text(aes(label = round(AUC, 3)),
            vjust = -0.9, size = 3, fontface = "bold") +
  scale_color_manual(values = c("Logistic Regression" = "#1f77b4",
                                "XGBoost"             = "#d62728")) +
  scale_x_continuous(breaks = wf_df$val_year) +
  scale_y_continuous(limits = c(
    max(0.5, min(wf_long$AUC, na.rm = TRUE) - 0.03),
    min(1.0, max(wf_long$AUC, na.rm = TRUE) + 0.05)
  )) +
  labs(
    title    = "Walk-Forward CV: AUC by Validation Year",
    subtitle = "Each point = AUC on held-out year immediately following training window",
    x = "Validation Year", y = "AUC", color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor  = element_blank(),
        legend.position   = "bottom")
```

**Interpretation:** Both models show stable AUC across the five walk-forward folds, confirming that the patterns learned from 1992–2012 generalise across different temporal cutpoints rather than being specific to any single training window. XGBoost's walk-forward AUC (`r round(wf_mean_xgb, 3)` ± `r round(wf_sd_xgb, 3)`) sits above logistic regression's (`r round(wf_mean_lr, 3)` ± `r round(wf_sd_lr, 3)`) across folds, consistent with the test-set results. The held-out test-set AUC for logistic regression (`r round(logit_test_auc, 3)`) falls within one standard deviation of its walk-forward mean, confirming the primary split is representative and not anomalously lucky or unlucky. The same holds for XGBoost (`r round(xgb_test_auc, 3)` vs. walk-forward mean `r round(wf_mean_xgb, 3)`).

## 4.11 Per-Year Recall: Did XGBoost Hold Up in 2015?

AUC declined in 2015, but the operationally relevant question is whether XGBoost continued to flag a meaningful share of actual fire months in an extreme year. A model that catches 60% of fires in a record drought year is more valuable than one that maintains AUC by correctly classifying the easy majority class of non-fire months. Per-year recall directly measures this.

```{r per-year-recall, fig.cap="Figure 8: Per-year recall for XGBoost and logistic regression on the held-out test set (2013–2015), at threshold = 0.5. Each bar shows the fraction of actual fire months correctly flagged in that calendar year. The 2015 decline in AUC did not translate into a collapse in XGBoost's recall — fire-month detection held up in the record drought year.", fig.width=7, fig.height=4}
recall_by_year <- lapply(sort(unique(test_data$YEAR)), function(yr) {
  idx      <- test_data$YEAR == yr
  actual   <- as.integer(test_data$notable_fire[idx]) - 1
  pred_xgb <- ifelse(xgb_test_prob[idx] >= 0.5, 1, 0)
  pred_lr  <- ifelse(logit_test_prob[idx] >= 0.5, 1, 0)
  n_pos    <- sum(actual == 1)
  data.frame(
    Year    = yr,
    XGBoost = if (n_pos > 0) round(100 * sum(pred_xgb == 1 & actual == 1) / n_pos, 1) else NA,
    Logistic = if (n_pos > 0) round(100 * sum(pred_lr  == 1 & actual == 1) / n_pos, 1) else NA
  )
})
recall_yr <- do.call(rbind, recall_by_year)

recall_yr_long <- tidyr::pivot_longer(recall_yr, cols = c(XGBoost, Logistic),
                                      names_to = "Model", values_to = "Recall_pct")

ggplot(recall_yr_long, aes(x = factor(Year), y = Recall_pct, fill = Model)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = paste0(Recall_pct, "%")),
            position = position_dodge(width = 0.6),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("XGBoost" = "#d62728", "Logistic" = "#1f77b4")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1),
                     limits = c(0, 100), expand = c(0, 0.02)) +
  labs(title    = "Per-Year Recall: XGBoost vs. Logistic Regression",
       subtitle = "Fraction of actual fire months correctly flagged | threshold = 0.5",
       x = "Year", y = "Recall (%)", fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "bottom")
```

**Interpretation:** XGBoost's recall held up across all three test years. The 2015 AUC decline did not correspond to a collapse in fire-month detection — the model continued flagging actual fire months above the 0.5 threshold at a meaningful rate even in the record drought year. This confirms the central argument: AUC and recall answer different questions. In 2015, XGBoost's probability rankings became less clean as drought spread widely and more non-fire months received elevated scores — but the model still correctly identified the highest-risk county-months. Logistic regression's recall remains near-zero in every year regardless of AUC, driven entirely by the absence of imbalance correction.

---

# Section 5: Summary

```{r summary-numbers, include=FALSE}
logit_auc_val <- round(as.numeric(logit_test_auc), 3)
xgb_auc_val   <- round(as.numeric(xgb_test_auc), 3)
logit_recall  <- logit_metrics["Recall"]
xgb_recall    <- xgb_metrics["Recall"]
```

The EDA established three structural patterns that shaped every modeling decision: wildfire incidence peaks sharply in the July–September window (with the cross-year average peaking in September, reflecting the western fire season extending into fall), is heavily concentrated in the western United States (~24× higher average monthly rates than the Northeast), and clusters temporally — a county that burned last month is substantially more likely to burn again, reflecting the persistence of the drought and heat conditions that cause large fires rather than reignition of the same ground.

On test AUC, XGBoost and logistic regression are close — `r round(as.numeric(xgb_test_auc), 3)` vs. `r round(as.numeric(logit_test_auc), 3)`. If AUC were the only lens, one might conclude the models perform similarly. But AUC treats a false positive and a false negative as equivalent errors, averaging performance over the full distribution of thresholds and both outcome classes. In wildfire forecasting, these errors are not equivalent. A false negative — failing to flag a county-month that burns — means no resources are pre-positioned and the fire response begins from scratch. A false positive — flagging a county-month that does not burn — triggers unnecessary pre-positioning, a recoverable logistical cost. Under this asymmetry, recall is the appropriate primary criterion: it measures directly how many dangerous months the model caught. On recall, the models are not close. XGBoost correctly flags `r round(100*xgb_recall, 1)`% of actual fire months against logistic regression's `r round(100*logit_recall, 1)`% — the difference between flagging approximately `r round(100*xgb_recall, 1)`% of dangerous months and fewer than one in ten. This gap is a consequence of deliberate imbalance correction via `scale_pos_weight`, not superior discrimination, which is why the AUC numbers remain similar while the recall numbers diverge sharply. The two metrics answer different questions: AUC measures how well a model can rank observations; recall measures whether it catches the ones that matter.

The primary limitations are feature scope and spatial structure. This model includes three nClimDiv climate variables (PDSI, temperature, precipitation) at the climate division level, but finer-grained covariates — county-level vapor pressure deficit, wind speed, satellite-derived fuel moisture — are known physical drivers that remain absent and would likely push both AUC and recall higher. Adjacent counties share climate and fire weather in ways the current county-independence assumption ignores, meaning standard errors are likely underestimated and effective sample size smaller than the observation count implies. The per-year AUC decline in 2015 — a record drought year — is consistent with the known limitation of historically-trained models in tail events, and further motivates recall-first evaluation: a model that catches more fires in an extreme year may be operationally superior even if its AUC falls. The time-based split (train 1992–2012, test 2013–2015), with 2012 reserved as a clean validation fold for XGBoost early stopping, ensures all reported metrics reflect genuine out-of-sample generalisation with no test set leakage. Walk-forward cross-validation (Section 4.10) confirms that both models' AUC is stable across training windows, validating the primary split as representative. A natural extension is **stacked generalisation** — feeding both models' predicted probabilities into a meta-learner that learns when to trust each, capturing XGBoost's nonlinear strength in high-risk county-months while inheriting logistic regression's stability in the low-risk majority.

The figure below distils the core tradeoff of the entire project into a single image.

```{r conclusion-figure, fig.cap="Figure 6: Recall breakdown for logistic regression and XGBoost on the held-out test set (2013–2015). Each bar represents the fate of actual fire months: the green segment shows the proportion correctly flagged (true positives) and the red segment shows the proportion missed (false negatives). XGBoost's substantially higher recall reflects deliberate calibration toward the lower-cost error type — a false alarm incurs recoverable pre-positioning costs, while a missed fire month results in an unprepared response to an event already in progress.", fig.width=8, fig.height=5.5}

recall_data <- data.frame(
  Model   = c("Logistic Regression", "Logistic Regression",
              "XGBoost",             "XGBoost"),
  Outcome = c("Correctly flagged (TP)", "Missed entirely (FN)",
              "Correctly flagged (TP)", "Missed entirely (FN)"),
  Pct     = c(round(100 * logit_metrics["Recall"], 1),
              round(100 - 100 * logit_metrics["Recall"], 1),
              round(100 * xgb_metrics["Recall"], 1),
              round(100 - 100 * xgb_metrics["Recall"], 1))
)

recall_data$Model   <- factor(recall_data$Model,
                              levels = c("Logistic Regression", "XGBoost"))
recall_data$Outcome <- factor(recall_data$Outcome,
                              levels = c("Missed entirely (FN)",
                                         "Correctly flagged (TP)"))

ggplot(recall_data, aes(x = Model, y = Pct, fill = Outcome)) +
  geom_col(width = 0.55, color = "white", linewidth = 0.4) +
  geom_text(
    aes(label = paste0(Pct, "%")),
    position = position_stack(vjust = 0.5),
    fontface = "bold", size = 5.5, color = "white"
  ) +
  scale_fill_manual(
    values = c("Missed entirely (FN)"   = "#b22222",
               "Correctly flagged (TP)" = "#2a7d2a"),
    name = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1),
                     limits = c(0, 100), expand = c(0, 0)) +
  labs(
    title    = "What happens to actual fire months?",
    subtitle = "Of every 100 county-months that experienced a notable wildfire, how many did each model catch?",
    x = NULL, y = "% of actual fire months"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "bottom",
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )
```

The figure above reduces the central finding to its most direct form. Of every 100 county-months that actually experienced a notable wildfire, logistic regression flagged `r round(100 * logit_metrics["Recall"], 1)`% and missed `r round(100 - 100 * logit_metrics["Recall"], 1)`%; XGBoost flagged `r round(100 * xgb_metrics["Recall"], 1)`% and missed `r round(100 - 100 * xgb_metrics["Recall"], 1)`%. The difference is not primarily a reflection of superior pattern recognition — it is a consequence of deliberately calibrating XGBoost toward the error type that carries lower operational cost. A false alarm triggers unnecessary resource pre-positioning; a missed fire triggers an unprepared response to an event already in progress. Where those costs are asymmetric, the threshold for prediction should be asymmetric too. That choice belongs to the agency, not the model — what the model provides is the full distribution of available tradeoffs across every possible operating point.

---

*Report generated using R Markdown. All code is available via the "Code" toggles above.*
